% -*- Mode: LaTeX -*-

\vspace*{50pt}
{\huge\bf Acknowledgments}\par
\vskip 40pt

The process of designing and implementing CLIM and writing the CLIM
specification has been a long and sometimes bumpy one.  The first thanks have
to go to the people and organizations who allowed the people working on CLIM
to continue to do so, despite the obstacles.  In particular, Mark Son-Bell of
ILA; Bob Laddaga and Saiid Zarrabian at Symbolics (and before or with them,
Ken Sinclair, Rick Karash, and Mark Graffam); and Jim Vietch and Hanoch Eiron
of Franz, Inc.  have struggled with various and sometimes conflicting business
interests through the whole process.

The attendees of the CLIM coalition meetings and those who contributed to CLIM
along the way include all of those listed on the title page, but also include
Bob Kerns, John Irwin, Paul Weineke and JonL White, and Kevin Males.  Other
contributors include Mike McMahon, Paul Robertson, John Hotchkiss, Dave Lowry,
Mike Greenwald, Dave Schatsky, and Glenn Adams.  Barry Margolin and Richard
Billington, acting for the ALU, contributed useful comments as well.

Sonya Keene, Ann Hathaway, and Ellen Golden all deserve credit for improving
the quality of this specification (in part by authoring the original CLIM
documentation), as does Mark Nahabedian.

Kent Pitman provided much useful advice on the mechanics of writing a
specification, and this document improved significantly because of that.

Last but not least are the people who have continued to use CLIM for the
duration of this lengthy process.  The people at BBN (Bruce Roberts and his
group) have been a continuing source of valuable information and suggestions.
Also, Markus Fischer of Symbolics GmbH had to deal with more than his fair
share of problems, the result of using CLIM in novel ways.
