% -*- Mode: LaTeX; Package: CLIM-USER -*-

\chapter {Changes from CLIM 1.0}
\label {changes}

This appendix lists the incompatible changes from CLIM 1.0 (and CLIM 0.9 for the
API related to the windowing substrate and gadgets), and the rationale for those
changes.  They are listed on a chapter-by-chapter basis.

When the items say that a compatibility stub will be provided, this does not
mean that this compatibility needs to be part of CLIM itself.  It could be
provided by a small compatibility package that defines stubs that translate from
the old behavior to the new behavior at compile-time or run-time, or by some
sort of conversion utility, or both.  In the first case, compiler warnings
should be generated to indicate that an obsolete form is being used.

\issue {SWM} {There are still lots of things from the windowing part, and the
frames, panes, and gadgets chapters that need to be included here.}

\paragraph {Regions}

\begin{itemize}
\item \cl{point-position*} has been renamed to \cl{point-position}, since the
term ``position'' unambiguously refers to an $(x,y)$ coordinate pair.  A
compatibility function will be provided.

\item \cl{region-contains-point*-p} has been renamed to
\cl{region-contains-position-p}, since the term ``position'' unambiguously
refers to an $(x,y)$ coordinate pair.  A compatibility function will be
provided.

\item The use of \cl{region-set-function} has been deprecated in favor of using
the three classes \cl{standard-region-union}, \cl{standard-region-intersection},
and \cl{standard-region-difference}, in keeping with the spirit of CLOS.
\cl{region-set-function} will be provided as a compatibility function.
\end{itemize}


\paragraph {Bounding Rectangles}

\begin{itemize} 
\item \cl{with-bounding-rectangle*} used to have optional \arg{max-x} and
\arg{max-y} arguments.  They are now required.

\item The function \cl{bounding-rectangle-set-edges} has been removed, since
bounding rectangles have been made immutable.  There is no replacement for it.

\item \cl{bounding-rectangle-position*} has been renamed to
\cl{bounding-rectangle-position}, since the term ``position'' unambiguously
refers to an $(x,y)$ coordinate pair.  A compatibility function will be
provided.

\item The functions \cl{bounding-rectangle-left}, \cl{bounding-rectangle-top},
\cl{bounding-rectangle-right}, and \cl{bounding-rectangle-bottom} have been
replaced by \cl{bounding-rectangle-min-x}, \cl{bounding-rectangle-min-y},
\cl{bounding-rectangle-max-x}, and \cl{bounding-rectangle-max-y}.  This is
because left, top, right, and bottom are ill-specified.  Compatibility functions
will be provided.
\end{itemize}


\paragraph {Affine Transformations}

\begin{itemize}
\item The function \cl{make-3-point-transformation} has had its argument list
changed from {\it (point-1 point-1-image point-2 point-2-image point-3 point-3-image)}
to {\it (point-1 point-2 point-3 point-1-image point-2-image point-3-image)}.
This was done because the original argument list did not group together inputs
and output, which was confusing.

\item The function \cl{make-3-point-transformation*} has had its argument list
changed from {\it (x1 y1 x1-image y1-image x2 y2 x2-image y2-image x3 y3 x3-image y3-image)}
to {\it (x1 y1 x2 y2 x3 y3 x1-image y1-image x2-image y2-image x3-image y3-image)}.
This was done because the original argument list did not group together inputs
and output, which was confusing.

\item \cl{compose-scaling-transformation}, \cl{compose-translation-transformation}, and
\cl{compose-rotation-transformation} have been replaced by the six functions
\cl{compose-translation-with-transformation}, \cl{compose-scaling-with-transformation},
\cl{compose-rotation-with-transformation}, \cl{compose-transformation-with-translation},
\cl{compose-transformation-with-scaling}, and \cl{compose-transformation-with-rotation}.
This was done because the six functions implement all of the optimized useful
cases of composition of transformations, and new names are required for all six.
Compatibility functions will be provided for the three CLIM 1.0 functions.

\item \cl{transform-point*} and \cl{untransform-point*} have been renamed to
\cl{transform-position} and \cl{untransform-position}.  Compatibility functions
will be provided.
\end{itemize}


\paragraph {Properties of Sheets}


\paragraph {Sheet Protocols}


\paragraph {Ports, Grafts, and Mirrored Sheets}


\paragraph {Text Styles}

\begin{itemize}
\item The macros \cl{with-text-style}, \cl{with-text-family},
\cl{with-text-face}, and \cl{with-text-size} have been changed to take the
\arg{medium} argument first and the text style (or family, face, or size)
argument second.  This was done in order to be consistent with all of the other
macros that take a \arg{medium} argument as the first argument.  Compatibility
code will be provided that attempts to detect the old syntax and massages it
into the new syntax, although it will probably not be able to detect all cases.

\item \cl{add-text-style-mapping} has been replaced by \cl{(setf~text-style-mapping)}
to be consistent with Common Lisp conventions.  A compatibility function will be
provided. 
\end{itemize}


\paragraph {Drawing in Color}

\begin{itemize}
\item \cl{+foreground+} and \cl{+background+} have been renamed to
\cl{+foreground-ink+} and \cl{+background-ink+}, for consistency with
\cl{+flipping-ink+}.  Compatibility constants will be provided.

\item \cl{make-color-rgb} and \cl{make-color-ihs} have been renamed to
\cl{make-rgb-color} and \cl{make-ihs-color}, by popular demand.  Compatibility
functions will be provided.
\end{itemize}


\paragraph {Extended Stream Output}

\begin{itemize}
\item \cl{stream-cursor-position*} and \cl{stream-increment-cursor-position*}
have been renamed to \cl{stream-cursor-position} and
\cl{stream-increment-cursor-position}.  Compatibility functions will be
provided.

\item The function \cl{stream-set-cursor-position*} has been replaced by
\cl{(setf*~stream-cursor-position)} to be consistent with Common Lisp
conventions.  A compatibility function will be provided.

\item The function \cl{stream-vsp} has been replace by
\cl{stream-vertical-spacing}.  A compatibility function will be provided.

\item The macros \cl{with-end-of-line-action} and \cl{with-end-of-page-action}
have been changed to take the \arg{stream} argument first and the action
argument second.  This was done in order to be consistent with all of the other
macros that take a \arg{stream} argument as the first argument.  Compatibility
code will be provided that attempts to detect the old syntax and massages it
into the new syntax, although it will probably not be able to detect all cases.
\end{itemize}


\paragraph {Output Recording}

\begin{itemize}
\item The three protocol classes \cl{output-record}, \cl{output-record-element},
and \cl{displayed-output-record-element} have been replaced by the two classes
\cl{output-record} and \cl{displayed-output-record}.  The predicates for the
classes have been similarly changed.

\item \cl{output-record-position*} has been renamed to
\cl{output-record-position}.  A compatibility function will be provided.

\item The function \cl{output-record-set-position*} has been replaced by
\cl{(setf*~output-record-position)} to be consistent with Common Lisp
conventions.  A compatibility function will be provided.

\item The functions \cl{output-record-start-position*},
\cl{output-record-set-start-position*}, \cl{output-record-end-position*},
\cl{output-record-set-end-position*} have been replaced by
\cl{output-record-start-cursor-position}, \cl{(setf*~output-record-start-cursor-position)},
\cl{output-record-end-cursor-position}, \cl{(setf*~output-record-end-cursor-position)} to
better reflect their functionality.  Compatibility functions will be provided.

\item \cl{replay-1} has been renamed to \cl{replay-output-record}.

\item \cl{output-record-elements} and \cl{output-record-element-count} have been
renamed to \cl{output-record-children} and \cl{output-record-count}, since the
term ``element'' is no longer used when referring to output records.
Compatibility functions will be provided.

\item \cl{add-output-record-element} and \cl{delete-output-record-element} have
been renamed to \cl{add-output-record} and \cl{delete-output-record}, and the
argument order has been changed.  Compatibility functions will be provided.

\item \cl{map-over-output-record-elements-containing-point*} and
\cl{map-over-output-record-elements-overlapping-region} have been renamed to
\cl{map-over-output-records-containing-position} and
\cl{map-over-output-records-overlapping-region}.  Compatibility functions will
be provided.

\item \cl{linear-output-record} and \cl{coordinate-sorted-set-output-record}
have been renamed to \cl{standard-sequence-output-record} and
\cl{standard-tree-output-record}.

\item \cl{stream-draw-p} and \cl{stream-record-p} and their \cl{setf} functions
have been renamed to \cl{stream-drawing-p} and \cl{stream-recording-p} to better
reflect their functionality.  Compatibility functions will be provided.

\item \cl{output-recording-stream-output-record},
\cl{output-recording-stream-current-output-record-stack}, and
\cl{output-recording-stream-text-output-record} have been renamed to
\cl{stream-output-history}, \cl{stream-current-output-record}, and
\cl{stream-text-output-record}.  Compatibility functions will be provided.

\item \cl{add-output-record} has been renamed to \cl{stream-add-output-record}.
Because of the change to \cl{add-output-record-element} above, no compatibility
function can be provided.

\item \cl{close-current-text-output-record} has been renamed to
\cl{stream-close-text-output-record}.  A compatibility function will be
provided.

\item \cl{add-string-output-to-output-record} and
\cl{add-character-output-to-output-record} have been renamed to
\cl{stream-add-string-output} and \cl{stream-add-character-output}.
Compatibility functions will be provided.

\item \cl{with-output-recording-options} has had its \cl{:draw-p} and
\cl{:record-p} keyword arguments changed to \cl{:draw} and \cl{:record} to
conform to Common Lisp naming conventions.  Compatibility code will be provided.
\end{itemize}


\paragraph {Table Formatting}

\begin{itemize}
\item The \cl{:inter-column-spacing}, \cl{:inter-row-spacing}, and
\cl{:multiple-columns-inter-column-spacing} options to \cl{formatting-table}
have been renamed to \cl{:x-spacing}, \cl{:y-spacing}, and
\cl{:multiple-columns-x-spacing} in order to be consistent with the pane
options.  Compatibility options will be provided.

\item The \cl{:minimum-width} and \cl{:minimum-height} options to
\cl{formatting-cell} have been renamed to \cl{:min-width} and \cl{:min-height}
in order to be consistent with the pane options.  Compatibility options will be
provided.

\item The \cl{:inter-column-spacing} and \cl{:inter-row-spacing} options to
\cl{formatting-item-list} and \cl{format-items} have been renamed to
\cl{:x-spacing} and \cl{:y-spacing} in order to be consistent with the pane
options.  Compatibility options will be provided.

\item The \cl{:no-initial-spacing} option to \cl{formatting-item-list} and
\cl{format-items} has been renamed to \cl{:initial-spacing}, because
inverted-sense flags are too hard to keep straight.  The default for
\cl{:no-initial-spacing} was \term{true}, therefore the default for
\cl{:initial-spacing} is \term{false}.  Compatibility options will be
provided.
\end{itemize}


\paragraph {Graph Formatting}

\begin{itemize}
\item The function \cl{format-graph-from-root} has been renamed to
\cl{format-graph-from-roots}, since it now takes a sequence of root objects,
rather than a single root object.  The function \cl{format-graph-from-root} will
remain as a compatibility function that takes a single root object.
\end{itemize}


\paragraph {Incremental Redisplay}

\begin{itemize}
\item \cl{redisplay-1} has been renamed to \cl{redisplay-output-record}.
\end{itemize}


\paragraph {Extended Stream Input}

\begin{itemize}
\item \cl{stream-pointer-position*} has been renamed to
\cl{stream-pointer-position}.  A compatibility function will be provided.

\item The function \cl{stream-set-pointer-position*} has been replaced by
\cl{(setf*~stream-pointer-position)} to be consistent with Common Lisp
conventions.  A compatibility function will be provided.

\item All of the clause arglists for \cl{tracking-pointer} are specified with
{\tt\&key}, that is, they are named arguments rather than positional ones.
This should not cause any problems, except for the one case that the
\arg{character} argument to the \cl{:keyboard} clause has been renamed to
\arg{gesture}.

\item The function \cl{dragging-output-record} has been renamed to
\cl{drag-output-record} to be consistent with our naming conventions.  A
compatibility function will be provided.
\end{itemize}


\paragraph {Presentation Types}

\begin{itemize}
\item The argument list for \cl{with-output-as-presentation} has been changed to
make \arg{stream}, \arg{object}, and \arg{type} be required arguments instead of
keyword arguments.  This is because it is always necessary to supply those
arguments in order for \cl{with-output-as-presentation} to work.  Compatibility
code will be provided to support the old syntax.

\item The \cl{:activation-characters}, \cl{:additional-activation-characters},
\cl{:blip-characters}, and \cl{:additional-blip-characters} keyword arguments to
the \cl{accept} functions have been renamed to \cl{:activation-gestures},
\cl{:additional-activation-gestures}, \cl{:delimiter-gestures}, and
\cl{:additional-delimiter-gestures}.  Compatibility code will be provided to
support the old keyword arguments.

\item The arglists for presentation translators and their documentation and
tester components have been changed to take a single positional \arg{object}
argument and a list of named (keyword) arguments.  Except for translators that
omit the \arg{object} argument or have it in other than the initial position of
the arglist, this will not pose a problem.  This change can be detected.

\item The \arg{frame} argument to \cl{find-presentation-translators} has
been changed to be a \arg{command-table} argument.  A check at run-time
can detect when a frame is supplied to
\cl{find-presentation-translators} instead of a command table.

\item The \cl{:shift-mask} keyword argument to \cl{test-presentation-translator},
\cl{find-applicable-translators}, \cl{presentation-matches-context-type}, and
\cl{find-innermost-applicable-presentation} has been renamed to
\cl{:modifier-state} in order to be consistent with the device event terminology.
Compatibility code will be provided to support the old keyword.

\item \cl{define-gesture-name} is completely different from CLIM 1.1.  There
will be no compatibility code provided to support the old version of
\cl{define-gesture-name}.

\item \cl{dialog-view} and \cl{+dialog-view+} have been renamed to
\cl{textual-dialog-view} and \cl{+textual-dialog-view+} in order to accurately
reflect what they are.  Likewise, \cl{menu-view} and \cl{+menu-view+} have been
renamed to \cl{textual-menu-view} and \cl{+textual-menu-view+}.  Compatibility
classes and constants will be provided.
\end{itemize}


\paragraph {Input Editing and Completion Facilities}

\begin{itemize}
\item \cl{*activation-characters*}, \cl{*standard-activation-characters*},
\cl{with-activation-characters}, and \cl{activation-character-p} have been
renamed to \cl{*activation-gestures*}, \cl{*standard-activation-gestures*},
\cl{with-activation-gestures}, and \cl{activation-gesture-p}.  Compatibility
functions will remain for \cl{with-activation-characters} and
\cl{activation-character-p}, but since the variables were not previously
documented, no compatibility will be provided for them.

\item \cl{*blip-characters*}, \cl{with-blip-characters}, and
\cl{blip-character-p} have been renamed to \cl{*delimiter-gestures*},
\cl{with-delimiter-gestures}, and \cl{delimiter-gesture-p}.  Compatibility
functions will remain for \cl{with-blip-characters} and \cl{blip-character-p},
but since \cl{*blip-characters*} was not previously documented, no compatibility
will be provided.

\item \cl{*abort-characters*} has been renamed to \cl{*abort-gestures*}.

\item \cl{*completion-characters*}, \cl{*help-characters*}, and
\cl{*possibilities-characters*} have been renamed to \cl{*completion-gestures*},
\cl{*help-gestures*}, and \cl{*possibilities-gestures*}.

\item Input editing streams no longer use the interactive stream class.
Instead, interactive streams are defined to be any stream that can potentially
support input editing, and the class \cl{input-editing-stream} now refers to
input editor streams.

\item \cl{input-editor-buffer}, \cl{input-position}, \cl{insertion-pointer}, and
\cl{rescanning-p} have been renamed to \cl{stream-input-buffer},
\cl{stream-scan-pointer}, \cl{stream-insertion-pointer}, and
\cl{stream-rescanning-p}.  Compatibility functions will be provided.
\end{itemize}


\paragraph {Menus}

\begin{itemize}
\item The \cl{:inter-column-spacing} and \cl{:inter-row-spacing} options to
\cl{menu-choose} have been renamed to \cl{:x-spacing} and \cl{:y-spacing} in
order to be consistent with the pane options.  Compatibility options will be
provided.
\end{itemize}


\paragraph {Command Processing}

\begin{itemize}
\item The variable \cl{*unsupplied-argument*} has been renamed to
\cl{*unsupplied-argument-marker*} in keeping with its functionality, and to
match the new \cl{*numeric-argument-marker*}.  \cl{*unsupplied-argument*} will
be retained, but its use is deprecated.

\item The \cl{:inter-column-spacing} and \cl{:inter-row-spacing} options to
\cl{display-command-table-menu} have been renamed to \cl{:x-spacing} and
\cl{:y-spacing} in order to be consistent with the pane options.  Compatibility
options will be provided.

\item The \cl{:test} argument to the following functions has been removed, since
the use of gesture names makes it unnecessary:
\cl{add-command-to-command-table}, \cl{(add-keystroke-to-command-table}, and
\cl{remove-keystroke-from-command-table}.  The \cl{:keystroke-test} argument has
been removed from \cl{read-command} and \cl{read-command-using-keystrokes} for
the same reason.
\end{itemize}


\paragraph {Application Frames}

\begin{itemize}
\item The \cl{:root} argument has been removed from \cl{open-window-stream} and
\cl{make-application-frame}.

\item The \cl{:layout} option has been removed, and is replaced by the more
general \cl{:layouts} option.  A compatibility hook will be provided that
handles the old \cl{:layout} option.

\item The function \cl{set-frame-layout} has been replaced by
\cl{(setf~frame-current-layout)} to be consistent with Common Lisp conventions.
A compatibility function will be provided.

\item The function \cl{frame-top-level-window} has been renamed to
\cl{frame-top-level-sheet}.  A compatibility function will be provided.

\item \cl{command-enabled-p}, \cl{enable-command}, and \cl{disable-command}
have been replaced by \cl{command-enabled} and \cl{(setf~command-enabled)}.
Compatibility functions will be provided.

\item \cl{window-viewport-position*} has been renamed to
\cl{window-viewport-position}.  A compatibility function will be provided.

\item \cl{window-set-viewport-position*} has been replaced by
{(setf*~window-viewport-position)}.  A compatibility function will be provided.
\end{itemize}


\paragraph {Panes}

\begin{itemize}
\item \cl{realize-pane} and \cl{realize-pane-1} have been renamed to
\cl{make-pane} and \cl{make-pane-1}.  A compatibility function will be provided
for \cl{realize-pane}.

\item The pane options \cl{:hs}, \cl{:hs+}, \cl{:hs-}, \cl{:vs}, \cl{:vs+}, and
\cl{:vs-} have been replaced by the options \cl{:width}, \cl{:max-width},
\cl{:min-width}, \cl{:height}, \cl{:max-height}, and \cl{:min-height} to be more
perspicuous, and to conform the same options for the formatted output
facilities.  Compatibility options will be supplied.

\item The \cl{:nchars} and \cl{:nlines} pane options have been removed in favor
of an extended syntax to the \cl{:width} and \cl{:height} options.

\item The pane layout options \cl{:halign} and \cl{:valign} have been renamed to
\cl{:align-x} and \cl{:align-y} to conform with table formatting.  Compatibility
options will be supplied.

\item The pane layout options \cl{:hspace} and \cl{:vspace} have been renamed to
\cl{:x-spacing} and \cl{:y-spacing} to conform with table formatting.
Compatibility options will be supplied.

\item The term ``space req'' has been renamed to ``space requirement''.  All of
the functions with \cl{space-req} in their names have been renamed to have
\cl{space-requirement} instead.

\item \cl{make-space-requirement} no longer takes the \cl{:hs} and \cl{:vs}
arguments, {\it et al}.  It now takes \cl{:width} and \cl{:height}, {\it et al}.
\end{itemize}
