% -*- Mode: LaTeX; Package: CLIM-USER -*-

\chapter {Presentation Types}
\label {presentation-types}

\section {Overview of Presentation Types}

The core around which the CLIM application user interface model is built is the
concept of the application-defined user interface data type.  Each application
has its own set of semantically significant user interface entities; a CAD
program for designing circuits has its various kinds of components (gates,
resistors, and so on), while a database manager has its relations and field
types.  These entities have to be displayed to the user (possibly in more than
one displayed representation) and the user has to be able to interact with and
specify the entities via pointer gestures and keyboard input.  Frequently each
user interface entity has a corresponding Lisp data type (such as an
application-specific structure or CLOS class definition), but this is not always
the case.  The data representation for an interaction entity may be a primitive
Lisp data type.  In fact, it is possible for several different user interface
entities to use the same Lisp data type for their internal representation, for
example, building floor numbers and employee vacation day totals could both be
represented internally as integers.

CLIM provides a framework for defining the appearance and behavior of these user
interface entities via the \concept{presentation type} mechanism.  A
presentation type can be thought of as a CLOS class that has some additional
functionality pertaining to its roles in the user interface of an application.
By defining a presentation type the application programmer defines all of the
user interface components of the entity:

\begin{itemize}
\item Its displayed representation, textual or graphical

\item Textual representation, for user input via the keyboard

\item Pointer sensitivity, for user input via the pointer
\end{itemize}

In other words, by defining a presentation type, the application programmer
describes in one place all the information about an object necessary to display
it to the user and interact with the user for object input.

The set of presentation types forms a type lattice, an extension of the Common
Lisp CLOS type lattice.  When a new presentation type is defined as a subtype of
another presentation type it inherits all the attributes of the supertype except
those explicitly overridden in the definition.

\issue {SWM} {Describe what a presentation type is more exactly.  What is a
parameterized presentation type?  Why do we want them?  Why are they in a
lattice?  How do they relate to CL types and CLOS classes?  What exactly gets
inherited?}


\section {Presentations}

A \concept{presentation} is a special kind of output record that remembers not
only output, but the object associated with the output and the semantic type
associated with that object.

\issue {SWM} {Describe exactly what a presentation is.  What does it mean for
presentations to be nested?}

\Defprotoclass {presentation}

The protocol class that corresponds to a presentation.
\IfYouWantClass {a} {presentation} {presentation}

\Defpredicate {presentationp} {object}

Returns \term{true} if \arg{object} is a \term{presentation}, otherwise returns
\term{false}.

\Defclass {standard-presentation}

The instantiable output record class that represents presentations.
\cl{present} normally creates output records of this class.
\Mutable

\definitarg {:object}
\definitarg {:type}
\definitarg {:view}
\definitarg {:single-box}
\Definitarg {:modifier}

All presentation classes must handle these five initargs, which are used to
specify, respectively, the object, type, view, single-box, and modifier
components of a presentation.


\subsection {The Presentation Protocol}

The following functions comprise the presentation protocol.  All classes that
inherit from \cl{presentation} must implement methods for these generic
functions.

\Defgeneric {presentation-object} {presentation}

Returns the object associated with the \term{presentation} \arg{presentation}.

\Defgeneric {(setf presentation-object)} {object presentation}

Changes the object associated with the \term{presentation} \arg{presentation} to
\arg{object}.

\Defgeneric {presentation-type} {presentation}

Returns the presentation type associated with the \term{presentation}
\arg{presentation}.

\Defgeneric {(setf presentation-type)} {type presentation}

Changes the object associated with the \term{presentation} \arg{presentation} to
\arg{object}.

\Defgeneric {presentation-single-box} {presentation}

Returns the ``single box'' attribute of the \term{presentation}
\arg{presentation}, which controls how the presentation is highlighted and when
it is sensitive.  This will be one of four values:

\begin{itemize} 
\item \cl{nil} (the default)---if the pointer is pointing at a visible piece of
the output that was drawn as part of the presentation, then it is considered to
be pointing at the presentation.  The presentation is highlighted by
highlighting each visible part of the output that was drawn as part of the
presentation.

\item \cl{t}---if the pointer is inside the bounding rectangle of the
presentation, it is considered to be pointing at the presentation.  The
presentation is highlighted by drawing a thin border around the bounding
rectangle.

\item \cl{:position}---like \cl{t} for determining whether the pointer is
pointing at the presentation, but like \cl{nil} for highlighting.

\item \cl{:highlighting}---like \cl{nil} for determining whether the pointer is
pointing at the presentation, but like \cl{t} for highlighting.
\end{itemize}

\Defgeneric {(setf presentation-single-box)} {single-box presentation}

Changes the ``single box'' attribute of the \term{presentation}
\arg{presentation} to \arg{single-box}.

\Defgeneric {presentation-modifier} {presentation}

Returns the ``modifier'' associated with the \term{presentation}
\arg{presentation}.  The modifier is some sort of object that describes how the
presentation object might be modified.  For example, it might be a function of
one argument (the new value) that can be called in order to store a new value
for \arg{object} after a user somehow ``edits'' the presentation.


\section {Presentation Types}

The type associated with a presentation is specified with a
\concept{presentation type specifier}, an object matching one of the following
three patterns: \\
\begin{tabular}{l}
  \arg{name} \\
  \cl{(\arg{name} \arg{parameters...})} \\
  \cl{((\arg{name} \arg{parameters...}) \arg{options...})}
\end{tabular}

Note that \arg{name} can be either a symbol that names a presentation type or a
CLOS class object (but not a \cl{built-in-class} object), in order to support
anonymous CLOS classes.

The \arg{parameters} ``parameterize'' the type, just as in a Common Lisp type
specifier.  The function \cl{presentation-typep} uses the parameters to check
object membership in a type.  Adding parameters to a presentation type specifier
produces a subtype, which contains some, but not necessarily all, of the objects
that are members of the unparameterized type.  Thus the parameters can turn off
the sensitivity of some presentations that would otherwise be sensitive.

% We decided to ignore the fact that this use of ``parameters'' conflicts with
% Common Lisp terminology for functions, which would call them ``arguments''.

The \arg{options} are alternating keywords and values that affect the use or
appearance of the presentation, but not its semantic meaning.  The \arg{options}
have no effect on presentation sensitivity.  (A programmer could choose to make
a tester in a translator examine options, but this is not standard practice.)
The standard option \cl{:description} is accepted by all types; if it is a
non-\cl{nil} value, then the value must be a string that describes the type and
overrides the description supplied by the type's definition.

Every presentation type is associated with a CLOS class.  If \arg{name} is a
class object or the name of a class, and that class is not a
\cl{built-in-class}, that class is the associated class.  Otherwise,
\cl{define-presentation-type} defines a class with metaclass
\cl{presentation-type-class} and superclasses determined by the presentation
type definition.  This class is not named \arg{name}, since that could interfere
with built-in Common Lisp types such as \cl{and}, \cl{member}, and \cl{integer}.
\cl{class-name} of this class returns a list \cl{(presentation-type \arg{name})}.
\cl{presentation-type-class} is a subclass of \cl{standard-class}.

Implementations are permitted to require programmers to evaluate the
\cl{defclass} form first in the case when the same name is used in both a
\cl{defclass} and a \cl{define-presentation-type}.

Every CLOS class (except for built-in classes) is a presentation type, as is its
name.  If it has not been defined with \cl{define-presentation-type}, it allows
no parameters and no options.

\concept{Presentation type inheritance} is used both to inherit methods (``what
parser should be used for this type?''), and to establish the semantics for the
type (``what objects are sensitive in this input context?'').  Inheritance of
methods is the same as in CLOS and thus depends only on the type name, not on
the parameters and options.

During presentation method combination, presentation type inheritance arranges
to translate the parameters of a subtype into a new set of parameters for its
supertype, and translates the options of the subtype into a new set of options
for the supertype.


\subsection {Defining Presentation Types}

\Defmacro {define-presentation-type} {name parameters
                                      \key options inherit-from description history
                                           parameters-are-types} 

Defines a presentation type whose name is the symbol or class \arg{name} and
whose parameters are specified by the lambda-list \arg{parameters}.  These
parameters are visible within \arg{inherit-from} and within the methods created
with \cl{define-presentation-method}.  For example, the parameters are used by
\cl{presentation-typep} and \cl{presentation-subtypep} methods to refine their
tests for type inclusion.

\arg{options} is a list of option specifiers.  It defaults to \cl{nil}.  An
option specifier is either a symbol or a list (\arg{symbol} \optional
\arg{default} \arg{supplied-p} \arg{presentation-type} \arg{accept-options}),
where \arg{symbol}, \arg{default}, and \arg{supplied-p} are as in a normal
lambda-list.  If \arg{presentation-type} and \arg{accept-options} are present,
they specify how to accept a new value for this option from the user.
\arg{symbol} can also be specified in the (\arg{keyword} \arg{variable}) form
allowed for Common Lisp lambda lists.  \arg{symbol} is a variable that is
visible within \arg{inherit-from} and within most of the methods created with
\cl{define-presentation-method}.  The keyword corresponding to \arg{symbol} can
be used as an option in the third form of a presentation type specifier.  An
option specifier for the standard option \cl{:description} is automatically
added to \arg{options} if an option with that keyword is not present, however it
does not produce a visible variable binding.

Unsupplied optional or keyword parameters default to \cl{*} (as in \cl{deftype})
if no default is specified in \arg{parameters}.  Unsupplied options default to
\cl{nil} if no default is specified in \arg{options}.

\arg{inherit-from} is a form that evaluates to a presentation type specifier for
another type from which the new type inherits.  \arg{inherit-from} can access
the parameter variables bound by the \arg{parameters} lambda list and the option
variables specified by \arg{options}.  If \arg{name} is or names a CLOS class
(other than a \cl{built-in-class}), then \arg{inherit-from} must specify the
class's direct superclasses (using \cl{and} to specify multiple inheritance).
It is useful to do this when you want to parameterize previously defined CLOS
classes.

If \arg{inherit-from} is unsupplied, it defaults as follows:  If \arg{name} is
or names a CLOS class, then the type inherits from the presentation type
corresponding to the direct superclasses of that CLOS class (using \cl{and} to
specify multiple inheritance).  Otherwise, the type named by \arg{name} inherits
from \cl{standard-object}.

\arg{description} is a string or \cl{nil}.  This should be the term for an
instance for the type being defined.  If it is \cl{nil} or unsupplied, a
description is automatically generated; it will be a ``prettied up'' version of
the type name, for example, \cl{small-integer} would become \cl{"small
integer"}.  You can also write a \cl{describe-presentation-type} presentation
method.  \arg{description} is implemented by the default
\cl{describe-presentation-type} method, so \arg{description} only works in
presentation types where that default method is not shadowed.

\arg{history} can be \cl{t} (the default), which means this type has its own
history of previous inputs, \cl{nil}, which means this type keeps no history,
or the name of another presentation type, whose history is shared by this
type.  More complex histories can be specified by writing a
\cl{presentation-type-history} presentation method.

\issue {SWM} {What is a presentation type history?  Should they be exposed?}

If the boolean \arg{parameters-are-types} is \term{true}, this means that the
parameters to the presentation type are themselves presentation types.  If they
are not presentation types, \arg{parameters-are-types} should be supplied as
\term{false}.  Types such as \cl{and}, \cl{or}, and \cl{sequence} will specify
this as \term{true}.

Every presentation type must define or inherit presentation methods for
\cl{accept} and \cl{present} if the type is going to be used for input and
output.  For presentation types that are only going to be used for input via the
pointer, the \cl{accept} need not be defined.

If a presentation type has \arg{parameters}, it must define presentation methods
for \cl{presentation-typep} and \cl{presentation-subtypep} that handle the
parameters, or inherit appropriate presentation methods.  In many cases it
should also define presentation methods for \cl{describe-presentation-type} and
\cl{presentation-type-specifier-p}.

There are certain restrictions on the \arg{inherit-from} form, to allow it to be
analyzed at compile time.  The form must be a simple substitution of parameters
and options into positions in a fixed framework.  It cannot involve conditionals
or computations that depend on valid values for the parameters or options; for
example, it cannot require parameter values to be numbers.  It cannot depend on
the dynamic or lexical environment.  The form will be evaluated at compile time
with uninterned symbols used as dummy values for the parameters and options.  In
the type specifier produced by evaluating the form, the type name must be a
constant that names a type, the type parameters cannot derive from options of
the type being defined, and the type options cannot derive from parameters of
the type being defined.  All presentation types mentioned must be already
defined.  \cl{and} can be used for multiple inheritance, but \cl{or}, \cl{not},
and \cl{satisfies} cannot be used.

None of the arguments, except \arg{inherit-from}, is evaluated.


\subsection {Presentation Type Abbreviations}

\Defmacro {define-presentation-type-abbreviation} {name parameters equivalent-type 
                                                   \key options}

\arg{name}, \arg{parameters}, and \arg{options} are as in
\cl{define-presentation-type}.  This defines a presentation type that is an
\concept{abbreviation} for the presentation type \arg{equivalent-type}.
Presentation type abbreviations can only be used in places where this
specification explicitly permits them.  In such places, \arg{equivalent-type}
and \term{abbreviation} are exactly equivalent and can be used interchangeably.

\arg{name} must be a symbol and must not be the name of a CLOS class.

The \arg{equivalent-type} form might be evaluated at compile time if
presentation type abbreviations are expanded by compiler optimizers.  Unlike
\arg{inherit-from}, \arg{equivalent-type} can perform arbitrary computations and
is not called with dummy parameter and option values.  The type specifier
produced by evaluating \arg{equivalent-type} can be a real presentation type or
another abbreviation.  If the type specifier doesn't include the standard option
\cl{:description}, the option is automatically copied from the abbreviation to
its expansion.

Note that you cannot define any presentation methods on a presentation type
abbreviation.  If you need methods, use \cl{define-presentation-type} instead.

\cl{define-presentation-type-abbreviation} is used to name a commonly used
cliche.  For example, a presentation type to read an octal integer might be
defined as
\begin{verbatim}
(define-presentation-type-abbreviation octal-integer (&optional low high) 
    `((integer ,low ,high) :base 8 :description "octal integer"))
\end{verbatim}

None of the arguments, except \arg{equivalent-type}, are evaluated.


\Defun {expand-presentation-type-abbreviation-1} {type \optional env}

If the \term{presentation type specifier} \arg{type} is a presentation type
abbreviation, or is an \cl{and}, \cl{or}, \cl{sequence}, or
\cl{sequence-enumerated} that contains a presentation type abbreviation, then
this expands the type abbreviation once, and returns two values, the expansion
and \cl{t}.  If \arg{type} is not a presentation type abbreviation, then the
values \arg{type} and \cl{nil} are returned.

\arg{env} is a macro-expansion environment, as for \cl{macroexpand}.

\Defun {expand-presentation-type-abbreviation} {type \optional env}

\cl{expand-presentation-type-abbreviation} is like
\cl{expand-presentation-type-abbreviation-1}, except that \arg{type} is
repeatedly expanded until all presentation type abbreviations have been removed.


\subsection {Presentation Methods}

Presentation methods inherit and combine in the same way as ordinary CLOS
methods.  The reason presentation methods are not exactly the same as ordinary
CLOS methods revolves around the \arg{type} argument.  The parameter specializer
for \arg{type} is handled in a special way, and presentation method inheritance
``massages'' the type parameters and options seen by each method.  For example,
consider three types \cl{int}, \cl{rrat}, and \cl{num} defined as follows:

\issue {SWM} {How are massaged arguments passed along?  Right now, we pass along
those parameters of the same name, and no others.}

\begin{verbatim}
(define-presentation-type int (low high)
  :inherit-from `(rrat ,high ,low))

(define-presentation-method presentation-typep :around (object (type int))
  (and (call-next-method)
       (integerp object)
       (<= low object high)))

(define-presentation-type rrat (high low)
  :inherit-from `num)

(define-presentation-method presentation-typep :around (object (type rrat))
  (and (call-next-method)
       (rationalp object)
       (<= low object high)))

(define-presentation-type num ())

(define-presentation-method presentation-typep (object (type num))
  (numberp object))
\end{verbatim}

If the user were to evaluate the form \cl{(presentation-typep X '(int 1 5))},
then the type parameters will be \cl{(1 5)} in the \cl{presentation-typep}
method for \cl{int}, \cl{(5 1)} in the method for \cl{rrat}, and \cl{nil} in the
method for \cl{num}.  The value for \arg{type} will be or \cl{((int 1 5))} in
each of the methods.


\Defmacro {define-presentation-generic-function} {generic-function-name 
                                                  presentation-function-name
                                                  lambda-list \rest options}

Defines a generic function that will be used for presentation methods.
\arg{generic-function-name} is a symbol that names the generic function that
will be used internally by CLIM for the individual methods,
\arg{presentation-function-name} is a symbol that names the function that
programmers will call to invoke the method, and \arg{lambda-list} and
\arg{options} are as for \cl{defgeneric}.

There are some ``special'' arguments in \arg{lambda-list} that are known about
by the presentation type system.  The first argument in \arg{lambda-list} must
be either \cl{type-key} or \cl{type-class}; this argument is used by CLIM to
implement method dispatching.  The second argument may be \cl{parameters},
meaning that, when the method is invoked, the type parameters will be passed to
it.  The third argument may be \cl{options}, meaning that, when the method is
invoked, the type options will be passed to it.  Finally, an argument named
\cl{type} must be included in \arg{lambda-list}; when the method is called,
\arg{type} argument will be bound to the presentation type specifier.

For example, the \cl{accept} presentation generic function might be defined as
follows:
\begin{verbatim}
(define-presentation-generic-function present-method present
  (type-key parameters options object type stream view
   &key acceptably for-context-type))
\end{verbatim}

None of the arguments are evaluated.


\Defmacro {define-presentation-method} {name qualifiers* specialized-lambda-list \body body}

Defines a presentation method for the function named \arg{name} on the
presentation type named in \arg{specialized-lambda-list}.
\arg{specialized-lambda-list} is a CLOS specialized lambda list for the method,
and its contents varies depending on what \arg{name} is.  \arg{qualifiers*} is
zero or more of the usual CLOS method qualifier symbols.
\cl{define-presentation-method} must support at least \cl{standard} method
combination (and therefore the \cl{:before}, \cl{:after}, and \cl{:around}
method qualifiers).  Some CLIM implementations may support other method
combination types, but this is not required.

\arg{body} defines the body of the method.  \arg{body} may have zero or more
declarations as its first forms.

All presentation methods have an argument named \arg{type} that must be
specialized with the name of a presentation type.  The value of \arg{type} is a
presentation type specifier, which can be for a subtype that inherited the
method.

All presentation methods except \cl{presentation-subtypep} have lexical access
to the parameters from the presentation type specifier.  Presentation methods
for the functions \cl{accept}, \cl{present}, \cl{describe-presentation-type},
\cl{presentation-type-specifier-p}, and \cl{accept-present-default} also have
lexical access to the options from the presentation type specifier.


\Defmacro {define-default-presentation-method} {name qualifiers* specialized-lambda-list
                                                \body body}

Like \cl{define-presentation-method}, except that it is used to define a default
method that will be used only if there are no more specific methods.

\Defmacro {funcall-presentation-generic-function} {presentation-function-name \rest arguments}

Calls the presentation generic function named by \arg{presentation-function-name}
on the arguments \arg{arguments}.  \arg{arguments} must match the arguments
specified by the \cl{define-presentation-generic-function} that was used to define
the presentation generic function, excluding the \cl{type-key}, \cl{type-class},
\cl{parameters}, and \cl{options} arguments, which are filled in by CLIM.

\cl{funcall-presentation-generic-function} is analogous to \cl{funcall}.

The \arg{presentation-function-name} argument is not evaluated.

For example, to call the \cl{present} presentation generic function, one might
use the following:
\begin{verbatim}
(funcall-presentation-generic-function present
  object presentation-type stream view)
\end{verbatim}


\Defmacro {apply-presentation-generic-function} {presentation-function-name \rest arguments}

Like \cl{funcall-presentation-generic-function}, except that
\cl{apply-presentation-generic-function} is analogous to \cl{apply}.

The \arg{presentation-function-name} argument is not evaluated.


Here is a list of all of the standard presentation methods and their specialized
lambda lists.  For the meaning of the arguments to each presentation method,
refer to the description of the function that calls that method.

For all of the presentation methods, the \arg{type} will always be specialized.
For those methods that take a \arg{view} argument, implementors and programmers
may specialize it as well.  The other arguments are not typically specialized.

\def\Defpresmeth #1 #2 {\Dodocf {#1} {#2} {Presentation~Method}}

\Defpresmeth {present} {object type stream view \key acceptably for-context-type} 

The \cl{present} presentation method is responsible for displaying the
representation of \arg{object} having \term{presentation type} \arg{type} for a
particular \term{view} \arg{view}.  The method's caller takes care of creating
the presentation, the method simply displays the content of the presentation.

The \cl{present} method can specialize on the \arg{view} argument in order to
define more than one view of the data.  For example, a spreadsheet program might
define a presentation type for revenue, which can be displayed either as a
number or a bar of a certain length in a bar graph.  Typically, at least one
canonical view should be defined for a presentation type, for example, the
\cl{present} method for the \cl{textual-view} view must be defined if the
programmer wants to allow objects of that type to be displayed textually.

{\bf Implementation note:} the actual argument list to the \cl{present} method is
\\
\arg{(type-key parameters options object type stream view \key acceptably for-context-type)}
\\
\arg{type-key} is the object that is used to cause the appropriate methods to be
selected (an instance of the class that corresponds to the presentation type
\arg{type}.).
\arg{parameters} and \arg{options} are the parameters and
options for the type on which the current method is specialized.  
% except in the slow, non-metaobject-protocol-based implementation where
% \arg{parameters} and \arg{options} are just decoded from \arg{type}
% and code inserted in the method body develops the parameters and
% options for the type on which the current method is specialized.  
The other arguments are gotten from the arguments of the same name in \cl{present}.


{\bf Implementation note:} the actual generic function of the \cl{present}
method is an internal generic function, not the function whose name is
\cl{present}.  Similar internal generic functions are used for all presentation
methods.


\Defpresmeth {accept} {type stream view \key default default-type}

The \cl{accept} method is responsible for ``parsing'' the representation of the
\term{presentation type} \arg{type} for a particular \term{view} \arg{view}.
The \cl{accept} method must return a single value, the object that was
``parsed'', or two values, the object and its type (a presentation type
specifier).  The method's caller takes care of establishing the input context,
defaulting, prompting, and input editing.

The \cl{accept} method can specialize on the \arg{view} argument in order to
define more than one input view for the data.  The \cl{accept} method for the
\cl{textual-view} view must be defined if the programmer wants to allow objects
of that type to entered via the keyboard.

Note that \cl{accept} presentation methods can call \cl{accept} recursively.  In
this case, the programmer should be careful to specify \cl{nil} for \cl{:prompt}
and \cl{:display-default} unless recursive prompting is really desired.

{\bf Implementation note:} the actual argument list to the \cl{accept} method is
\\
\arg{(type-key parameters options type stream view \key default default-type)}


\Defpresmeth {describe-presentation-type} {type stream plural-count}

The \cl{describe-presentation-type} method is responsible for textually
describing the \term{presentation type} \arg{type}.  \arg{stream} is a stream,
and will not be \cl{nil} as it can be for the \cl{describe-presentation-type}
function.

{\bf Implementation note:} the actual argument list to the 
\cl{describe-presentation-type} method is
\\
\arg{(type-key parameters options type stream plural-count)} 


\Defpresmeth {presentation-type-specifier-p} {type}

The \cl{presentation-type-specifier-p} method is responsible for checking the
validity of the parameters and options for the \term{presentation type}
\arg{type}.  The default method returns \cl{t}.

{\bf Implementation note:} the actual argument list to the
\cl{presentation-type-specifier-p} method is
\\
\arg{(type-key parameters options type)}


\Defpresmeth {presentation-typep} {object type}

The \cl{presentation-typep} method is called when the \cl{presentation-typep}
function requires type-specific knowledge.  If the type name in the
\term{presentation type} \arg{type} is a CLOS class or names a CLOS class, the
method is called only if \arg{object} is a member of the class and \arg{type}
contains parameters, and the method simply tests whether \arg{object} is a
member of the subtype specified by the parameters.  For non-class types, the
method is always called.

{\bf Implementation note:} the actual argument list to the \cl{presentation-typep}
method is
\\
\arg{(type-key parameters object type)}


\Defpresmeth {presentation-subtypep} {type putative-supertype}

\cl{presentation-subtypep} walks the type lattice (using
\cl{map-over-presentation-supertypes}) to determine whethe or not the
\term{presentation type} \arg{type} is a subtype of the\term{presentation type}
\arg{putative-supertype}, without looking at the type parameters.  When a
supertype of \arg{type} has been found whose name is the same as the name of
\arg{putative-supertype}, then the \cl{subtypep} method for that type is called
in order to resolve the question by looking at the type parameters (that is, if
the \cl{subtypep} method is called, \arg{type} and \arg{putative-supertype} are
guaranteed to be the same type, differing only in their parameters).  If
\arg{putative-supertype} is never found during the type walk, then
\cl{presentation-subtypep} will never call the \cl{presentation-subtypep}
presentation method for \arg{putative-supertype}.

Unlike all other presentation methods, \cl{presentation-subtypep} receives a
\arg{type} argument that has been translated to the presentation type for which
the method is specialized; \arg{type} is never a subtype.  The method is only
called if \arg{putative-supertype} has parameters and the two presentation type
specifiers do not have equal parameters.  The method must return the two values
that \cl{presentation-subtypep} returns.

Since \cl{presentation-subtypep} takes two type arguments, the parameters are
not lexically available as variables in the body of a presentation method.

{\bf Implementation note:} the actual argument list to the
\cl{presentation-subtypep} method is
\\
\arg{(type-key type putative-supertype)}


\Defpresmeth {map-over-presentation-type-supertypes} {function type}

This method is called in order to apply \arg{function} to the superclasses of
the \term{presentation type} \arg{type}.

{\bf Implementation note:} the actual argument list to the
\cl{map-over-presentation-type-supertypes} method is
\\
\arg{(type-class function type)}


\Defpresmeth {accept-present-default} {type stream view default default-supplied-p
                                       present-p query-identifier}

The \cl{accept-present-default} method is called when \cl{accept} turns into
\cl{present} inside of \cl{accepting-values}.  The default method calls
\cl{present} or \cl{describe-presentation-type} depending on whether
\arg{default-supplied-p} is \term{true} or \term{false}, respectively.

The boolean \arg{default-supplied-p} will be \term{true} only in the case when
the \cl{:default} option was explicitly supplied in the call to \cl{accept} that
invoked \cl{accept-present-default}.

{\bf Implementation note:} the actual argument list to the
\cl{accept-present-default} method is
\\
\arg{(type-key parameters options type stream view default default-supplied-p present-p query-identifier)}


\Defpresmeth {presentation-type-history} {type}

This method is responsible for returning a history object for the
\term{presentation type} \arg{type}.

{\bf Implementation note:} the actual argument list to the
\cl{presentation-type-history} method is
\\
\arg{(type-key parameters type)}


\Defpresmeth {presentation-default-preprocessor} {default type \key default-type}

This method is responsible for taking the object \arg{default}, and coercing it
to match the \term{presentation type} \arg{type} (which is the type being
accepted) and \arg{default-type} (which is the presentation type of
\arg{default}).  This is useful when you want to change the default gotten from
the presentation type's history so that it conforms to parameters or options in
\arg{type} and \arg{default-type}.)  The method must return two values, the new
object to be used as the default, and a new presentation type, which should be
at least as specific as \arg{type}.

{\bf Implementation note:} the actual argument list to the
\cl{presentation-default-preprocessor} method is
\\
\arg{(type-key parameters default type \key default-type)}


\Defpresmeth {presentation-refined-position-test} {type record x y}

This method used to definitively answer hit detection queries for a
presentation, that is, determining that the point $(x,y)$ is contained within
the output record \arg{record}.  Its contract is exactly the same as for
\cl{output-record-refined-position-test}, except that it is intended to
specialize on the presentation type \arg{type}.

{\bf Implementation note:} the actual argument list to the
\cl{presentation-refined-position-test} method is
\\
\arg{(type-key parameters options type record x y)}


\Defpresmeth {highlight-presentation} {type record stream state}

This method is responsible for drawing a highlighting box around the
\term{presentation} \arg{record} on the \term{output recording stream}
\arg{stream}.  \arg{state} will be either \cl{:highlight} or \cl{:unhighlight}.

{\bf Implementation note:} the actual argument list to the
\cl{highlight-presentation} method is
\\
\arg{(type-key parameters options type record stream state)}



\subsection {Presentation Type Functions}

\Defun {describe-presentation-type} {type \optional stream plural-count}

Describes the \term{presentation type specifier} \arg{type} on the \term{stream}
\arg{stream}, which defaults to \cl{*standard-output*}.  If \arg{stream} is
\cl{nil}, a string containing the description is returned.  \arg{plural-count}
is either \cl{nil} (meaning that the description should be the singular form of
the name), \cl{t} (meaning that the description should the plural form of the
name), or an integer greater than zero (the number of items to be described).
The default is \cl{1}.

\arg{type} can be a presentation type abbreviation.


\Defun {presentation-type-parameters} {type-name \optional env}

Returns a lambda-list, the parameters specified when the presentation type or
presentation type abbreviation whose name is \arg{type-name} was defined.
\arg{type-name} is a symbol or a class.  \arg{env} is a macro-expansion
environment, as in \cl{find-class}.


\Defun {presentation-type-options} {type-name \optional env}

Returns the list of options specified when the presentation type or presentation
type abbreviation whose name is \arg{type-name} was defined.  This does not
include the standard options unless the presentation-type definition mentioned
them explicitly.  \arg{type-name} is a symbol or a class.  \arg{env} is a
macro-expansion environment, as in \cl{find-class}.


\Defmacro {with-presentation-type-decoded} {(name-var \optional parameters-var options-var) 
                                            type
                                            \body body}

The specified variables are bound to the components of the presentation type
specifier produced by evaluating \arg{type}, the forms in \arg{body} are
executed, and the values of the last form are returned.  \arg{name-var}, if
non-\cl{nil}, is bound to the presentation type name.  \arg{parameters-var}, if
non-\cl{nil}, is bound to a list of the parameters.  \arg{options-var}, if
non-\cl{nil}, is bound to a list of the options.  When supplied, \arg{name-var},
\arg{parameters-var}, and \arg{options-var} must be symbols.

The \arg{name-var}, \arg{parameters-var}, and \arg{options-var} arguments are
not evaluated.  \arg{body} may have zero or more declarations as its first
forms. 

\Defun {presentation-type-name} {type}

Returns the presentation type name of the presentation type specifier \arg{type}.
This function is provided as a convenience.  It could be implemented with the
following code:

\begin{verbatim}
(defun presentation-type-name (type)
  (with-presentation-type-decoded (name) type
    name))
\end{verbatim}

\Defmacro {with-presentation-type-parameters} {(type-name type) \body body}

Variables with the same name as each parameter in the definition of the
presentation type are bound to the parameter values in \arg{type}, if present,
or else to the defaults specified in the definition of the presentation type.
The forms in \arg{body} are executed in the scope of these variables and the
values of the last form are returned.

The value of the form \arg{type} must be a presentation type specifier whose
name is \arg{type-name}.  The \arg{type-name} and \arg{type} arguments are not
evaluated.  \arg{body} may have zero or more declarations as its first forms.

\Defmacro {with-presentation-type-options} {(type-name type) \body body}

Variables with the same name as each option in the definition of the
presentation type are bound to the option values in \arg{type}, if present, or
else to the defaults specified in the definition of the presentation type.  The
forms in \arg{body} are executed in the scope of these variables and the values
of the last form are returned.

The value of the form \arg{type} must be a presentation type specifier whose
name is \arg{type-name}.  The \arg{type-name} and \arg{type} arguments are not
evaluated.  \arg{body} may have zero or more declarations as its first forms.


\Defun {presentation-type-specifier-p} {object}

Returns \term{true} if \arg{object} is a valid \term{presentation type
specifier}, otherwise returns \term{false}.


\Defun {presentation-typep} {object type}

Returns \term{true} if \arg{object} is of the presentation type specified by the
\term{presentation type specifier} \arg{type}, otherwise returns \term{false}.

\arg{type} may not be a presentation type abbreviation.

This is analogous to the Common Lisp \cl{typep} function.


\Defun {presentation-type-of} {object}

Returns a presentation type of which \arg{object} is a member.
\cl{presentation-type-of} returns the most specific presentation type that can
be conveniently computed and is likely to be useful to the programmer.  This is
often the class name of the class of the object.

If \cl{presentation-type-of} cannot determine the presentation type of the
object, it may return either \cl{expression} or \cl{t}.

This is analogous to the Common Lisp \cl{typep} function.


\Defun {presentation-subtypep} {type putative-supertype}

Answers the question ``is the type specified by the \term{presentation type
specifier} \arg{type} a subtype of the type specified by the \term{presentation
type specifier} \arg{putative-supertype}?''.  \cl{presentation-subtypep} returns
two values, \arg{subtypep} and \arg{known-p}.  When \arg{known-p} is
\term{true}, \arg{subtypep} can be either \term{true} (meaning that \arg{type}
is definitely a subtype of \arg{putative-supertype}) or \term{false} (meaning
that \arg{type} is definitely not a subtype of \arg{putative-supertype}).  When
\arg{known-p} is \term{false}, then \arg{subtypep} must also be \term{false};
this means that the answer cannot reliably be determined.

\arg{type} may not be a presentation type abbreviation.

This is analogous to the Common Lisp \cl{subtypep} function.


\Defun {map-over-presentation-type-supertypes} {function type}

Calls the function \arg{function} on the \term{presentation type specifier}
\arg{type} and each of its supertypes.  \arg{function} is called with two
arguments, the name of a type and a presentation type specifier for that type
with the parameters and options filled in.  \arg{function} has dynamic extent;
its two arguments are permitted to have dynamic extent.  The traversal of the
type lattice is done in the order specified by the CLOS class precedence rules,
and visits each type in the lattice exactly once.


\Defun {presentation-type-direct-supertypes} {type}

Returns a sequence consisting of the names of all of the presentation types that
are direct supertypes of the \term{presentation type specifier} \arg{type}, or
\cl{nil} if \arg{type} has no supertypes.  The consequences of modifying the
returned sequence are unspecified.


\Defun {find-presentation-type-class} {name \optional (errorp \cl{t}) environment}

Returns the class corresponding to the presentation type named \arg{name}, which
must be a symbol or a class object.  \arg{errorp} and \arg{environment} are as
for \cl{find-class}.


\Defun {class-presentation-type-name} {class \optional environment}

Returns the presentation type name corresponding to the class \arg{class}.  This
is essentially the inverse of \cl{find-presentation-type-class}.
\arg{environment} is as for \cl{find-class}.


\Defun {default-describe-presentation-type} {description stream plural-count}

Performs the default actions for \cl{describe-presentation-type}, notably
pluralization and prepending an indefinite article if appropriate.
\arg{description} is a string or a symbol, typically the \cl{:description}
presentation type option or the \cl{:description} option to
\cl{define-presentation-type}.  \arg{plural-count} is as for
\cl{describe-presentation-type}.


\Defun {make-presentation-type-specifier} {type-name-and-parameters \rest options}

A convenient way to assemble a presentation type specifier with only non-default
options included.  This is only useful for abbreviation expanders, not for
\cl{:inherit-from}.  \arg{type-name-and-parameters} is a presentation type
specifier, which must be in the \cl{(\arg{type-name} \arg{parameters...})} form.
\arg{options} are alternating keywords and values that are added as options to
the presentation type specifier, except that if a value is equal to
\arg{type-name}'s default, that option is omitted, producing a more concise
presentation type specifier.


\section {Typed Output}

An application can specify that all output done within a certain dynamic extent
should be associated with a given Lisp object and be declared to be of a
specified presentation type.  The resulting output is saved in the window's
output history as a presentation.  Specifically, the presentation remembers the
output that was performed (by saving the associated output record), the Lisp
object associated with the output, and the presentation type specified at output
time.  The object can be any Lisp object.


\Defmacro {with-output-as-presentation} {(stream object type
                                          \key modifier single-box allow-sensitive-inferiors
                                               parent record-type \allow)
                                         \body body} 

The output of \arg{body} to the \term{extended output recording} \arg{stream} is
used to generate a presentation whose underlying object is \arg{object} and
whose presentation type is \arg{type}.  Each invocation of this macro results in
the creation of a presentation object in the stream's output history unless
output recording has been disabled or \cl{:allow-sensitive-inferiors nil} was
specified at a higher level, in which case the presentation object is not
inserted into the history.  \cl{with-output-as-presentation} returns the
presentation corresponding to the output.

The \arg{stream} argument is not evaluated, and must be a symbol that is bound
to an extended output stream or output recording stream.  If \arg{stream} is
\cl{t}, \cl{*standard-output*} is used. \arg{body} may have zero or more
declarations as its first forms.

\arg{type} may be a presentation type abbreviation.

\arg{modifier}, which defaults to \cl{nil}, is some sort of object that
describes how the presentation object might be modified.  For example, it might
be a function of one argument (the new value) that can be called in order to
store a new value for \arg{object} after a user somehow ``edits'' the
presentation.  \arg{modifier} must have indefinite extent.

\arg{single-box} is used to specify the \cl{presentation-single-box} component
of the resulting presentation.  It can take on the values described under
\cl{presentation-single-box}.

When the boolean \arg{allow-sensitive-inferiors} is \term{false}, nested calls
to \cl{present} or \cl{with-output-as-presentation} inside this one will not
generate presentations.  The default is \term{true}.

\arg{parent} specifies what output record should serve as the parent for the
newly created presentation.  If unspecified, \cl{stream-current-output-record}
of \arg{stream} will be used as the parent.

\arg{record-type} specifies the class of the presentation output record to be
created.  It defaults to \cl{standard-presentation}.  This argument should only
be supplied by a programmer if there is a new class of output record that
supports the updating output record protocol.

All arguments of this macro are evaluated.

For example,
\begin{verbatim}
(with-output-as-presentation (stream #p"foo" 'pathname)
  (princ "FOO" stream))
\end{verbatim}


\Defun {present} {object \optional type
                  \key stream view modifier acceptably for-context-type
                       single-box allow-sensitive-inferiors sensitive record-type}

The \arg{object} of \term{presentation type} \arg{type} is presented to the
\term{extended output stream} \arg{stream} (which defaults to
\cl{*standard-output*}), using the type's \cl{present} method for the supplied
\term{view} \arg{view}.  \arg{type} is a presentation type specifier, and can be
an abbreviation.  It defaults to \cl{(presentation-type-of \arg{object})}.  The
other arguments and overall behavior of \cl{present} are as for
\cl{stream-present}.

The returned value of \cl{present} is the presentation object that contains the
output corresponding to the object.

\cl{present} must be implemented by first expanding any presentation type
abbreviations (\arg{type} and \arg{for-context-type}), and then calling
\cl{stream-present} on \arg{stream}, \arg{object}, \arg{type}, and the remaining
keyword arguments, which are described below.

\Defgeneric {stream-present} {stream object type
                              \key view modifier acceptably for-context-type
                                   single-box allow-sensitive-inferiors sensitive 
                                   record-type}

\cl{stream-present} is the per-stream implementation of \cl{present}, analogous
to the relationship between \cl{write-char} and \cl{stream-write-char}.  All
extended output streams and output recording streams must implement a method for
\cl{stream-present}.  The default method (on
\cl{standard-extended-output-stream}) implements the following behavior.

The object \arg{object} of type \arg{type} is presented to the \term{stream}
\arg{stream} by calling the type's \cl{present} method for the supplied
\term{view} \arg{view}.  The returned value is the presentation containing the
output corresponding to the object.

\arg{type} is a presentation type specifier.  \arg{view} is a view object that
defaults to \cl{stream-default-view} of \arg{stream}.

\arg{for-context-type} is a presentation type specifier that is passed to the
\cl{present} method for \arg{type}, which can use it to tailor how the object
will be presented.  \arg{for-context-type} defaults to \arg{type}.

\arg{modifier}, \arg{single-box}, \arg{allow-sensitive-inferiors}, and
\arg{record-type} are the same as for \cl{with-output-as-presentation}.

\arg{acceptably} defaults to \cl{nil}, which requests the \cl{present} method to
produce text designed to be read by human beings.  If \arg{acceptably} is
\cl{t}, it requests the \cl{present} method to produce text that is recognized
by the \cl{accept} method for \arg{for-context-type}.  This makes no difference
to most presentation types.

The boolean \arg{sensitive} defaults to \term{true}.  If it is \term{false}, no
presentation is produced.


\Defun {present-to-string} {object \optional type
                            \key view acceptably for-context-type string index}

Same as \cl{present} inside \cl{with-output-to-string}.  If \arg{string} is
supplied, it must be a string with a fill pointer.  When \arg{index} is
supplied, it is used as an index into \arg{string}.  \arg{view},
\arg{acceptably}, and \arg{for-context-type} are as for \cl{present}.

The first returned value is the string.  When \arg{string} is supplied, a second
value is returned, the updated \arg{index}.


\section {Context-dependent (Typed) Input}

Associating semantics with output is only half of the user interface equation.
The presentation type system also supports the input side of the user
interaction.  When an application wishes to solicit from the user input of a
particular presentation type, it establishes an \concept{input context} for that
type.  CLIM will then automatically allow the user to satisfy the input request
by pointing at a visible presentation of the requested type (or a valid subtype)
and pressing a pointer button.  Only the presentations that ``match'' the input
context will be ``sensitive'' (that is, highlighted when the pointer is moved
over them) and accepted as input, thus the presentation-based input mechanism
supports \concept{context-dependent input}.

\issue {SWM} {What exactly is an input context?  What does it mean for them to
be nested?}

\Defvar {*input-context*}

The current input context.  This will be a list, each element of which
corresponds to a single call to \cl{with-input-context}.  The first element of
the list represents the context established by the most recent call to
\cl{with-input-context}, and the last element represents the context established
by the least recent call to \cl{with-input-context}.

The exact format of the elements in the list is unspecified, but will typically
be a list of a presentation type and a tag that corresponds to the point in the
control structure of CLIM at which the input context was establish.
\cl{*input-context*} and the elements in it may have dynamic extent.

\Defun {input-context-type} {context-entry}

Given one element from \cl{*input-context*}, \arg{context-entry}, returns the
presentation type of the context entry.

\Defmacro {with-input-context} {(type \key override)
                                (\optional object-var type-var event-var options-var)
                                form
                                \body pointer-cases} 

Establishes an input context of \term{presentation type} \arg{type}; this must
be done by binding \cl{*input-context*} to reflect the new input context.  When
the boolean \arg{override} is \term{false} (the default), this invocation of
\cl{with-input-context} adds its context presentation type to the current
context.  In this way an application can solicit more than one type of input at
the same time.  When \arg{override} is \term{true}, it overrides the current
input context rather than nesting inside the current input context.

\arg{type} can be a presentation type abbreviation.

After establishing the new input context, \arg{form} is evaluated.  If no
pointer gestures are made by the user during the evaluation of \arg{form}, the
values of \arg{form} are returned.  Otherwise, one of the \arg{pointer-cases} is
executed (based on the presentation type of the object that was clicked on) and
the value of that is returned.  (See the descriptions of
\cl{call-presentation-menu} and \cl{throw-highlighted-presentation}.)
\arg{pointer-cases} is constructed like a \cl{typecase} statement clause list whose
keys are presentation types; the first clause whose key satisfies the condition
\cl{(presentation-subtypep \arg{type} \arg{key})} is the one that is chosen.

During the execution of one of the \arg{pointer-cases}, \arg{object-var} is
bound to the object that was clicked on (the first returned value from the
presentation translator that was invoked), \arg{type-var} is bound to its
presentation type (the second returned value from the translator), and
\arg{event-var} is bound to the pointer button event that was used.
\arg{options-var} is bound to any options that a presentation translator might
have returned (the third value from the translator), and will be either \cl{nil}
or a list of keyword-value pairs.  \arg{object-var}, \arg{type-var},
\arg{event-var}, and \arg{options-var} must all be symbols.

\arg{type}, \arg{stream}, and \arg{override} are evaluated, the others are not.

For example,
\begin{verbatim}
(with-input-context ('pathname)
                    (path)
     (read)
   (pathname
     (format t "~&The pathname ~A was clicked on." path)))
\end{verbatim}


\Defun {accept} {type
                 \key stream view
                      default default-type provide-default 
                      insert-default replace-input history active-p
                      prompt prompt-mode display-default query-identifier
                      activation-gestures additional-activation-gestures
                      delimiter-gestures additional-delimiter-gestures}

Requests input of type \arg{type} from the \term{stream} \arg{stream}, which
defaults to \cl{*standard-input*}.  \cl{accept} returns two values, the object
representing the input and its presentation type.  \arg{type} is a presentation
type specifier, and can be an abbreviation.  The other arguments and overall
behavior of \cl{accept} are as for \cl{accept-1}.

\cl{accept} must be implemented by first expanding any presentation type
abbreviations (\arg{type}, \arg{default-type}, and \arg{history}), handling the
interactions between the default, default type, and presentation history,
prompting the user by calling \cl{prompt-for-accept}, and then calling
\cl{stream-accept} on \arg{stream}, \arg{type}, and the remaining keyword
arguments.

\Defgeneric {stream-accept} {stream type
                             \key view 
                                  default default-type provide-default 
                                  insert-default replace-input history active-p
                                  prompt prompt-mode display-default query-identifier
                                  activation-gestures additional-activation-gestures
                                  delimiter-gestures additional-delimiter-gestures}

\cl{stream-accept} is the per-stream implementation of \cl{accept}, analogous to
the relationship between \cl{read-char} and \cl{stream-read-char}.  All extended
input streams must implement a method for \cl{stream-accept}.  The default
method (on \cl{standard-extended-input-stream}) simply calls \cl{accept-1}.

The arguments and overall behavior of \cl{stream-accept} are as for \cl{accept-1}.

{\bf Rationale:} the reason \cl{accept} is specified as a three-function
``trampoline'' is to allow close tailoring of the behavior of \cl{accept}.
\cl{accept} itself is the function that should be called by application
programmers.  CLIM implementors will specialize \cl{stream-accept} on a
per-stream basis.  (For example, the behavior of \cl{accepting-values} can be
implemented by creating a special class of stream that turns calls to
\cl{accept} into fields of a dialog.)  \cl{accept-1} is provided as a
convenient function for the \cl{stream-accept} methods to call when they require
the default behavior.

\Defun {accept-1} {stream type
                   \key view 
                        default default-type provide-default 
                        insert-default replace-input history active-p
                        prompt prompt-mode display-default query-identifier
                        activation-gestures additional-activation-gestures
                        delimiter-gestures additional-delimiter-gestures}

Requests input of type \arg{type} from the \term{stream} \arg{stream}.
\arg{type} must be a presentation type specifier.  \arg{view} is a view object
that defaults to \cl{stream-default-view} of \arg{stream}.  \cl{accept-1}
returns two values, the object representing the input and its presentation type.
(If \cl{frame-maintain-presentation-histories} is \term{true} for the current
frame, then the returned object is also pushed on to the presentation history
for that object.)

\cl{accept-1} establishes an input context via \cl{with-input-context}, and then
calls the \cl{accept} presentation method for \arg{type} and \arg{view}.  When
called on an interactive stream, \cl{accept} must allow input editing; see
Chapter~\ref{input-editing} for a discussion of input editing.  The call to
\cl{accept} will be terminated when the \cl{accept} method returns, or the user
clicks on a sensitive presentation.  The typing of an activation and delimiter
character is typically one way in which a call to an \cl{accept} method is
terminated.

A top-level \cl{accept} satisfied by keyboard input discards the terminating
keyboard gesture (which will be either a delimiter or an activation gesture).
A nested call to \cl{accept} leaves the terminating gesture unread.

If the user clicked on a matching presentation, \cl{accept-1} will insert
the object into the input buffer by calling \cl{presentation-replace-input} on
the object and type returned by the presentation translator, unless either the
boolean \arg{replace-input} is \term{false} or the presentation translator
returned an \cl{:echo} option of \term{false}.  \arg{replace-input} defaults to
\term{true}, but this default is overridden by the translator explicitly
returning an \cl{:echo} option of \term{false}.

If \arg{default} is supplied, then it and \arg{default-type} are returned as
values from \cl{accept-1} when the input is empty.  \arg{default-type}
must be a presentation type specifier.  If \arg{default} is not supplied and
\arg{provide-default} is \term{true} (the default is \term{false}), then the
default is determined by taking the most recent item from the presentation type
history specified by \arg{history}.  If \arg{insert-default} is \term{true} and
there is a default, the default will be inserted into the input stream by
calling \cl{presentation-replace-input}.

\arg{history} must be either \cl{nil}, meaning that no presentation type history
will be used, or a presentation type (or abbreviation) that names a history to
be used for the call to \cl{accept}.  \arg{history} defaults to \arg{type}.

\arg{prompt} can be \cl{t}, which prompts by describing the type, \cl{nil},
which suppresses prompting, or a string, which is displayed as a prompt (via
\cl{write-string}).  The default is \cl{t}, which produces ``Enter a
\arg{type}:'' in a top-level call to \cl{accept} or ``(\arg{type})'' in a nested
call to \cl{accept}.

If the boolean \arg{display-default} is \term{true}, the default is displayed
(if one was supplied).  If \arg{display-default} is \term{false}, the default is
not displayed.  \arg{display-default} defaults to \term{true} if \arg{prompt}
was provided, otherwise it defaults to \term{false}.

\arg{prompt-mode} can be \cl{:normal} (the default) or \cl{:raw}, which
suppresses putting a colon after the prompt and/or default in a top-level
\cl{accept} and suppresses putting parentheses around the prompt and/or default
in a nested \cl{accept}.

\arg{query-identifier} is used within \cl{accepting-values} to identify the
field within the dialog.  The \cl{active-p} argument (which defaults to \cl{t})
can be used to control whether a field within an \cl{accepting-values} is
active; when \term{false}, the field will not be active, that is, it will not be
available for input.  Some CLIM implementations will provide a visual cue that
the field is inactive, for instance, by ``graying out'' the field.

\arg{activation-gestures} is a list of gesture names that will override the
current activation gestures (which are stored in \cl{*activation-gestures*}).
Alternatively, \arg{additional-activation-gestures} can be supplied to add
activation gestures without overriding the current ones.  See
Chapter~\ref{input-editing} for a discussion of activation gestures.

\arg{delimiter-gestures} is a list of gesture names that will override the
current delimiter gestures (which are stored in \cl{*delimiter-gestures*}).
Alternatively, \arg{additional-delimiter-gestures} can be supplied to add
delimiter gestures without overriding the current ones.  See
Chapter~\ref{input-editing} for a discussion of delimiter gestures.


\Defun {accept-from-string} {type string \key view default default-type start end}

Like \cl{accept}, except that the input is taken from \arg{string}, starting at
the position specified by \arg{start} and ending at \arg{end}.  \arg{view},
\arg{default}, and \arg{default-type} are as for \cl{accept}.

\cl{accept-from-string} returns an object and a presentation type (as in
\cl{accept}), but also returns a third value, the index at which input
terminated.


\Defgeneric {prompt-for-accept} {stream type view \rest accept-args \key}

Called by \cl{accept} to prompt the user for input of \term{presentation type}
\arg{type} on the \term{stream} \arg{stream} for the \term{view} \arg{view}.
\arg{accept-args} are all of the keyword arguments supplied to \cl{accept}.  The
default method (on \cl{standard-extended-input-stream}) simply calls
\cl{prompt-for-accept-1}.

\Defun {prompt-for-accept-1} {stream type
                              \key default default-type display-default
                                   prompt prompt-mode
                              \allow}

Prompts the user for input of \term{presentation type} \arg{type} on the
\term{stream} \arg{stream}.

If the boolean \arg{display-default} is \term{true}, then the default is
displayed; otherwise, the default is not displayed.  When the default is being
displayed, \arg{default} and \arg{default-type} are the taken as the object and
presentation type of the default to display. \arg{display-default} defaults to
\term{true} if \arg{prompt} is non-\cl{nil}, otherwise it defaults to
\term{false}.

If \arg{prompt} is \cl{nil}, no prompt is displayed.  If it is a string, that
string is displayed as the prompt.  If \arg{prompt} is \cl{t} (the default), the
prompt is generated by calling \cl{describe-presentation-type} to produce a
prompt of the form ``Enter a \arg{type}:'' in a top-level call to \cl{accept},
or ``(\arg{type})'' in a nested call to \cl{accept}.

\arg{prompt-mode} can be \cl{:normal} (the default) or \cl{:raw}, which
suppresses putting a colon after the prompt and/or default in a top-level
\cl{accept} and suppresses putting parentheses around the prompt and/or default
in a nested \cl{accept}.


\section {Views}

\cl{accept} and \cl{present} methods can specialize on the \arg{view} argument
in order to define more than one view of the data.  For example, a spreadsheet
program might define a presentation type for quarterly earnings, which can be
displayed as a floating point number or as a bar of some length in a bar graph.
These two views might be implemented by specializing the view arguments for the
\cl{textual-view} class and the user-defined \cl{bar-graph-view} class.

\Defprotoclass {view}

The protocol class for view objects.
\IfYouWantClass {a} {view} {view}

All of the view classes are immutable.

\Defpredicate {viewp} {object}

Returns \term{true} if \arg{object} is a \term{view}, otherwise returns
\term{false}.

\Defclass {textual-view}

The instantiable class representing all textual views, a subclass of \cl{view}.
Presentation methods that apply to a textual view must only do textual input and
output (such as \cl{read-char} and \cl{write-string}).

\Defclass {textual-menu-view}

The instantiable class that represents the default view that is used inside
\cl{menu-choose} for frame managers that are not using a gadget-oriented look
and feel.  It is a subclass of \cl{textual-view}.

\Defclass {textual-dialog-view}

The instantiable class that represents the default view that is used inside
\cl{accepting-values} dialogs for frame managers that are not using a
gadget-oriented look and feel.  It is a subclass of \cl{textual-view}.

\Defclass {gadget-view}

The instantiable class representing all gadget views, a subclass of \cl{view}.

\Defclass {gadget-menu-view}

The instantiable class that represents the default view that is used inside
\cl{menu-choose} for frame managers that are using a gadget-oriented look and
feel.  It is a subclass of \cl{gadget-view}.

\Defclass {gadget-dialog-view}

The instantiable class that represents the default view that is used inside
\cl{accepting-values} dialogs for frame managers that are using a
gadget-oriented look and feel.  It is a subclass of \cl{gadget-view}.

\Defclass {pointer-documentation-view}

The instantiable class that represents the default view that is used when
computing pointer documentation.  It is a subclass of \cl{textual-view}.

\defconst {+textual-view+}
\defconst {+textual-menu-view+}
\defconst {+textual-dialog-view+}
\defconst {+gadget-view+}
\defconst {+gadget-menu-view+}
\defconst {+gadget-dialog-view+}
\Defconst {+pointer-documentation-view+}

These are objects of class \cl{textual-view}, \cl{textual-menu-view},
\cl{textual-dialog-view}, \cl{gadget-view}, \cl{gadget-menu-view},
\cl{gadget-dialog-view}, and \cl{pointer-documentation-view}, respectively.

\Defgeneric {stream-default-view} {stream}

Returns the default view for the extended stream \arg{stream}.  \cl{accept} and
\cl{present} get the default value for the \arg{view} argument from this.  All
extended input and output streams must implement a method for this generic
function.

\Defgeneric {(setf stream-default-view)} {view stream}

Changes the default view for \arg{stream} to the \term{view} \arg{view}.  All
extended input and output streams must implement a method for this generic
function.


\section {Presentation Translators}

CLIM provides a mechanism for \concept{translating} between types.  In other
words, within an input context for presentation type {\it A} the translator
mechanism allows a programmer to define a translation from presentations of some
other type {\it B} to objects that are of type {\it A}.

Note that the exact representation of a presentation translator has been left
explicitly unspecified.


\subsection {Defining Presentation Translators}

\Defmacro {define-presentation-translator} {name 
                                            (from-type to-type command-table
                                             \key gesture
                                                  tester tester-definitive
                                                  documentation pointer-documentation 
                                                  menu priority)
                                            arglist
                                            \body body}

Defines a presentation translator named \arg{name} that translates from objects
of type \arg{from-type} to objects of type \arg{to-type}.  \arg{from-type} and
\arg{to-type} are presentation type specifiers, but must not include any
presentation type options.  \arg{from-type} and \arg{to-type} may be
presentation type abbreviations.

\arg{command-table} is a \term{command table designator}.  The translator created
by this invocation of \cl{define-presentation-translator} will be stored in the
command table \arg{command-table}.

\arg{gesture} is a gesture name that names a pointer gesture (described in
Section~\ref{gesture-names}).  The body of the translator will be run only if
the translator is applicable and gesture used by the user matches the gesture
name in the translator.  (We will explain \concept{applicability}, or
\concept{matching}, in detail below.)  \arg{gesture} defaults to \cl{:select}.

\arg{tester} is either a function or a list of the form
\\
\arg{(tester-arglist . tester-body)}
\\
where \arg{tester-arglist} takes the same form as \arg{arglist} (see below), and
\arg{tester-body} is the body of the tester.  The tester must return either
\term{true} or \term{false}.  If it returns \term{false}, then the translator is
definitely not applicable.  If it returns \term{true}, then the translator might
be applicable, and the body of the translator might be run (if
\arg{tester-definitive} is \term{false}) in order to definitively decide if the
translator is applicable (this is described in more detail below).  If no tester
is supplied, CLIM supplies a tester that always returns \term{true}.

When the boolean \arg{tester-definitive} is \term{true}, the body of the
translator will never be run in order to decide if the translator is applicable,
that is, the tester is assumed to definitively decide whether the translator
applies.  The default for \arg{tester-definitive} is \term{false}.  When there
is no explicitly supplied tester, the tester supplied by CLIM is assumed to be
definitive.

Both \arg{documentation} and \arg{pointer-documentation} are objects that will
be used for documenting the translator.  \arg{pointer-documentation} will be
used to generate documentation for the pointer documentation window; the
documentation generated by \arg{pointer-documentation} should be very brief and
computing it should be very fast and preferably not cons.  \arg{documentation}
is used to generate such things as items in the \cl{:menu}-gesture menu.  If the
object is a string, the string itself will be used as the documentation.
Otherwise, the object must be the name of a function or a list of the form
\\
\arg{(doc-arglist . doc-body)}
\\
where \arg{doc-arglist} takes the same form as \arg{arglist}, but includes a
named (keyword) \arg{stream} argument as well (see below), and \arg{doc-body}
is the body of the documentation function.  The body of the documentation
function should write the documentation to \arg{stream}.  The default for
\arg{documentation} is \cl{nil}, meaning that there is no explicitly supplied
documentation; in this case, CLIM is free to generate the documentation in
other ways.  The default for \arg{pointer-documentation} is \arg{documentation}.

\arg{menu} must be \cl{t} or \cl{nil}.  When it is \cl{t}, the translator will
be included in the \cl{:menu}-gesture menu if it matches.  When it is \cl{nil},
the translator will not be included in the \cl{:menu}-gesture menu.  Other
non-\cl{nil} values are reserved for future extensions to allow multiple
presentation translator menus.

\arg{priority} is either \cl{nil} (the default, which corresponds to 0) or an
integer that represents the priority of the translator.  When there are several
translators that match for the same gesture, the one with the highest priority
is chosen.

\arg{arglist}, \arg{tester-arglist}, and \arg{doc-arglist} are each an argument
list that must ``match'' the following ``canonical'' argument list.
\\
\arg{(object \key presentation context-type frame event window x y)}
\\
In order to ``match'' the canonical argument list, there must be a single
positional argument that corresponds to the presentation's object, and
several named arguments that must match the canonical names above (using
\cl{string-equal} to do the comparison).

In the body of the translator (or the tester), the positional \arg{object}
argument will be bound to the presentation's object.  The named arguments
\arg{presentation} will be bound to the presentation that was clicked on,
\arg{context-type} will be bound to the presentation type of the context that
actually matched, \arg{frame} will be bound to the application frame that is
currently active (usually \cl{*application-frame*}), \arg{event} will be bound
to the pointer button event that the user used, \arg{window} will be bound to
the window stream from which the event came, and \arg{x} and \arg{y} will be
bound to the $x$ and $y$ positions within \arg{window} that the pointer was at
when the event occurred.  The special variable \cl{*input-context*} will be
bound to the current input context.  Note that, in many implementations
\arg{context-type} and \cl{*input-context*} will have dynamic extent, so
programmers should not store without first copying them.

\arg{body} is the body of the translator, and is run in the context of the
application.  \arg{body} may have zero or more declarations as its first forms.
It should return either one, two, or three values.  The first value is an object
which must be \cl{presentation-typep} of \arg{to-type}, and the second value is
a presentation type that must be \cl{presentation-subtypep} of \arg{to-type}.
The consequences are unspecified if the object is not \cl{presentation-typep} of
\arg{to-type} or the type is not \cl{presentation-subtypep} of \arg{to-type}.
The first two returned values of \arg{body} are used, in effect, as the returned
values for the call to \cl{accept} that established the matching input context.

The third value returned by \arg{body} must either be \cl{nil} or a list of
options (as keyword-value pairs) that will be interpreted by \cl{accept}.  The
only option defined so far is \cl{:echo}, whose value must be either \term{true}
(the default) or \term{false}.  If it is \term{true}, the object returned by the
translator will be ``echoed'' by \cl{accept}, which will use
\cl{presentation-replace-input} to insert the textual representation of the
object into the input buffer.  If it is \term{false}, the object will not be
echoed.

None of \cl{define-presentation-translator}'s arguments are evaluated.


\Defmacro {define-presentation-to-command-translator} {name 
                                                       (from-type command-name command-table
                                                        \key gesture tester
                                                             documentation pointer-documentation 
                                                             menu priority echo)
                                                       arglist
                                                       \body body}

This is similar to \cl{define-presentation-translator}, except that the
\arg{to-type} will be derived to be the command named by \arg{command-name} in
the command table \arg{command-table}.  \arg{command-name} is the name of the
command that this translator will translate to.

The \arg{echo} option is a boolean value (the default is \term{true}) that
indicates whether the command line should be echoed when a user invokes the
translator.

The other arguments to \cl{define-presentation-to-command-translator} are the
same as for \cl{define-presentation-translator}.  Note that the tester for
command translators is always assumed to be definitive, so there is no
\cl{:tester-definitive} option. The default for \arg{pointer-documentation} is
the string \arg{command-name} with dash characters replaced by spaces, and each
word capitalized (as in \cl{add-command-to-command-table}).

The body of the translator must return a list of the arguments to the command
named by \arg{command-name}.  \arg{body} is run in the context of the
application.  The returned value of the body, appended to the command name, are
eventually passed to \cl{execute-frame-command}.  \arg{body} may have zero or
more declarations as its first forms.

None of \cl{define-presentation-to-command-translator}'s arguments are evaluated.


\Defmacro {define-presentation-action} {name 
                                        (from-type to-type command-table 
                                         \key gesture tester
                                              documentation pointer-documentation
                                              menu priority)
                                        arglist
                                        \body body}

\cl{define-presentation-action} is similar to \cl{define-presentation-translator},
except that the body of the action is not intended to return a value, but should
instead side-effect some sort of application state.

A presentation action does not satisfy a request for input the way an ordinary
translator does.  Instead, an action is something that happens while waiting for
input.  After the action has been executed, the program continues to wait for
the same input that it was waiting for prior to executing the action.

The other arguments to \cl{define-presentation-action} are the same as for
\cl{define-presentation-translator}.  Note that the tester for presentation
actions is always assumed to be definitive.

None of \cl{define-presentation-action}'s arguments are evaluated.


\Defmacro {define-drag-and-drop-translator} {name
                                             (from-type to-type destination-type command-table 
                                              \key gesture tester
                                                   documentation pointer-documentation
                                                   menu priority
                                                   feedback highlighting)
                                             arglist
                                             \body body}

Defines a ``drag and drop'' (or ``direct manipulation'') translator named
\arg{name} that translates from objects of type \arg{from-type} to objects of
type \arg{to-type} when a ``from presentation'' is ``picked up'', ``dragged''
over, and ``dropped'' on to a ``to presentation'' having type
\arg{destination-type}.  \arg{from-type}, \arg{to-type}, and
\arg{destination-type} are presentation type specifiers, but must not include
any presentation type options.  \arg{from-type}, \arg{to-type} and
\arg{destination-type} may be presentation type abbreviations.

The interaction style used by these translators is that a user points to a
``from presentation'' with the pointer, picks it up by pressing a pointer button
matching \arg{gesture}, drags the ``from presentation'' to a ``to presentation''
by moving the pointer, and then drops the ``from presentation'' onto the ``to
presentation''.  The dropping might be accomplished by either releasing the
pointer button or clicking again, depending on the frame manager.  When the
pointer button is released, the translator whose \arg{destination-type} matches
the presentation type of the ``to presentation'' is chosen.  For example,
dragging a file to the TrashCan on a Macintosh could be implemented by a drag
and drop translator.

While the pointer is being dragged, the function specified by \arg{feedback} is
invoked to provide feedback to the user.  The function is called with eight
arguments: the application frame object, the ``from presentation'', the stream,
the initial $x$ and $y$ positions of the pointer, the current $x$ and $y$
positions of the pointer, and a feedback state (either \cl{:highlight} to draw
feedback, or \cl{:unhighlight} to erase it).  The feedback function is called to
draw some feedback the first time pointer moves, and is then called twice each
time the pointer moves thereafter (once to erase the previous feedback, and then
to draw the new feedback).  It is called a final time to erase the last feedback
when the pointer button is released.  \arg{feedback} defaults to
\cl{frame-drag-and-drop-feedback}.

When the ``from presentation'' is dragged over any other presentation that has a
direct manipulation translator, the function specified by \arg{highlighting} is
invoked to highlight that object.  The function is called with four arguments:
the application frame object, the ``to presentation'' to be highlighted or
unhighlighted, the stream, and a highlighting state (either \cl{:highlight} or
\cl{:unhighlight}).  \arg{highlighting} defaults to
\cl{frame-drag-and-drop-highlighting}.

Note that it is possible for there to be more than one drag and drop translator
that applies to the same from-type, to-type, and gesture.  In this case, the
exact translator that is chosen for use during the dragging phase is
unspecified.  If these translators have different feedback, highlighting,
documentation, or pointer documentation, the exact behavior is unspecified.

The other arguments to \cl{define-drag-and-drop-translator} are the same as for
\cl{define-presentation-translator}.


\subsection {Presentation Translator Functions}

\Defun {find-presentation-translators} {from-type to-type command-table}

Returns a list of all of the translators in the \term{command table}
\arg{command-table} that translate from \arg{from-type} to \arg{to-type},
without taking into account any type parameters or testers.  \arg{from-type} and
\arg{to-type} are presentation type specifiers, and must not be abbreviations.
\arg{frame} must be an application frame.

{\bf Implementation note:} Because \cl{find-presentation-translators} is called
during pointer sensitivity computations (that is, whenever the user mouses the
pointer around in any CLIM pane), it should cache its result in order to avoid
consing.  Therefore, the resulting list of translators should not be modified;
the consequences of doing so are unspecified.

{\bf Implementation note:} The ordering of the list of translators is left
unspecified, but implementations may find it convenient to return the list using
the ordering specified for \cl{find-applicable-translators}.


\Defun {test-presentation-translator} {translator presentation context-type frame window x y
                                       \key event modifier-state for-menu}

Returns \term{true} if the translator \arg{translator} applies to the
presentation \arg{presentation} in input context type \arg{context-type},
otherwise returns \term{false}.  (There is no \arg{from-type} argument because
it is derived from \arg{presentation}.)  \arg{x} and \arg{y} are the $x$ and $y$
positions of the pointer within the window stream \arg{window}.

\arg{event} and \arg{modifier-state} are a pointer button event and modifier
state (see \cl{event-modifier-key-state}), and are compared against the
translator's gesture.  \arg{event} defaults to \cl{nil}, and
\arg{modifier-state} defaults to 0, meaning that no modifier keys are held down.
Only one of \arg{event} or \arg{modifier-state} may be supplied; it is
unspecified what will happen if both are supplied.

If \arg{for-menu} is \term{true}, the comparison against \arg{event} and
\arg{modifier-state} is not done.

\arg{presentation}, \arg{context-type}, \arg{frame}, \arg{window}, \arg{x},
\arg{y}, and \arg{event} are passed along to the translator's tester if and
when the tester is called.

\cl{test-presentation-translator} is responsible for matching type parameters
and calling the translator's tester.  Under some circumstances,
\cl{test-presentation-translator} may also call the body of the translator to
ensure that its value matches \arg{to-type}.


\Defun {find-applicable-translators} {presentation input-context frame window x y
                                      \key event modifier-state for-menu fastp}

Returns a list that describes the translators that definitely apply to the
\term{presentation} \arg{presentation} in the input context \arg{input-context}.
Each element in the returned list is of the form
\\
\arg{(translator the-presentation context-type . rest)}
\\
where \arg{translator} is a presentation translator, \arg{the-presentation} is
the presentation that the translator applies to (and can be different from
\arg{presentation} due to nesting of presentations), \arg{context-type} is the
context type in which the translator applies, and \arg{rest} is other
unspecified data reserved for internal use by CLIM.  \arg{translator},
\arg{the-presentation}, and \arg{context-type} can be passed to such functions
as \cl{call-presentation-translator} and \cl{document-presentation-translator}.

Since input contexts can be nested, \cl{find-applicable-translators} must
iterate over all the contexts in \arg{input-context}.  \arg{window}, \arg{x},
and \arg{y} are as for \cl{test-presentation-translator}.  \arg{event} and
\arg{modifier-state} (which default to \cl{nil} and the current modifier state
for \arg{window}, respectively) are used to further restrict the set of
applicable translators.  (Only one of \arg{event} or \arg{modifier-state} may be
supplied; it is unspecified what will happen if both are supplied.)

Presentations can also be nested.  The ordering of the translators returned by
\cl{find-applicable-translators} is that translators matching inner contexts
should precede translators matching outer contexts, and, in the same input
context, inner presentations precede outer presentations.

When \arg{for-menu} is non-\cl{nil}, this matches the value of \arg{for-menu}
against the presentation's menu specification, and returns only those
translators that match.  \arg{event} and \arg{modifier-state} are disregarded in
this case.  \arg{for-menu} defaults to \cl{nil}.

When the boolean \arg{fastp} is \term{true}, \cl{find-applicable-translators}
will simply return \term{true} if there are any translators.  \arg{fastp}
defaults to \term{false}.

When \arg{fastp} is \term{false}, the list of translators returned by
\cl{find-applicable-translators} must be in order of their ``desirability'',
that is, translators having more specific from-types and/or higher priorities
must precede translators having less specific from-types and lower priorities.

The rules used for ordering the translators returned by
\cl{find-applicable-translators} are as follows (in order):

\begin{enumerate}
\item Translators with a higher ``high order'' priority precede translators with
a lower ``high order'' priority.  This allows programmers to set the priority of
a translator in such a way that it always precedes all other translators.

\item Translators with a more specific ``from type'' precede translators with a
less specific ``from type''.

\item Translators with a higher ``low order'' priority precede translators with
a lower ``low order'' priority.  This allows programmers to break ties between
translators that translate from the same type.

\item Translators from the current command table precede translators inherited
from superior command tables.
\end{enumerate}

{\bf Implementation note:} \cl{find-applicable-translators} could be implemented
by looping over \arg{input-context}, calling \cl{find-presentation-translators}
to generate all the translators, and then calling \cl{test-presentation-translator}
to filter out the ones that do not apply.  The consequences of modifying the
returned value are unspecified.  Note that the ordering of translators can be
done by \cl{find-presentation-translators}, provided that
\cl{find-applicable-translators} takes care to preserve this ordering.

\issue {SWM} {Describe and implement the \cl{class-nondisjoint-classes} idea.
Be very clear and precise about when the translator body gets run.}

\Defun {presentation-matches-context-type} {presentation context-type frame window x y
                                            \key event modifier-state}

Returns \term{true} if there are any translators that translate from the
\term{presentation} \arg{presentation}'s type to the input context type
\arg{context-type}, otherwise returns \term{false}.  (There is no
\arg{from-type} argument because it is derived from \arg{presentation}.)
\arg{frame}, \arg{window}, \arg{x}, \arg{y}, \arg{event}, and
\arg{modifier-state} are as for \cl{test-presentation-translator}.

If there are no applicable translators, \cl{presentation-matches-context-type}
will return \term{false}.


\Defun {call-presentation-translator} {translator presentation context-type
                                       frame event window x y}

Calls the function that implements the body of the translator \arg{translator}
on the \term{presentation} \arg{presentation}'s object, and passes
\arg{presentation}, \arg{context-type}, \arg{frame}, \arg{event}, \arg{window},
\arg{x}, and \arg{y} to the body of the translator as well.

The returned values are the same as the values returned by the body of the
translator, namely, the translated object and the translated type.


\Defun {document-presentation-translator} {translator presentation context-type
                                           frame event window x y
                                           \key (stream \cl{*standard-output*}) documentation-type}  

Computes the documentation string for the translator \arg{translator} and
outputs it to the stream \arg{stream}.  \arg{presentation}, \arg{context-type},
\arg{frame}, \arg{event}, \arg{window}, \arg{x}, and \arg{y} are as for
\cl{test-presentation-translator}.

\cl{documentation-type} must be either \cl{:normal} or \cl{:pointer}.  If it is
\cl{:normal}, the usual translator documentation function is called.  If it is
\cl{:pointer}, the translator's pointer documentation is called.


\Defun {call-presentation-menu} {presentation input-context frame window x y
                                 \key for-menu label}

Finds all the applicable translators for the \term{presentation}
\arg{presentation} in the input context \arg{input-context}, creates a menu that
contains all of the translators, and pops up the menu from which the user can
choose a translator.  After the translator is chosen, it is called with the
arguments supplied to \cl{call-presentation-menu} and the matching input context
that was established by \cl{with-input-context} is terminated.

\arg{window}, \arg{x}, \arg{y}, and \arg{event} are as for
\cl{find-applicable-translators}.  \arg{for-menu}, which defaults to \cl{t}, is
used to decide which of the applicable translators will go into the menu; only
those translators whose \cl{:menu} option matches \arg{menu} will be included.

\arg{label} is either a string to use as a label for the menu, or is \cl{nil}
(the default), meaning the menu will not be labelled.


\subsection {Finding Applicable Presentations}

\Defun {find-innermost-applicable-presentation} {input-context window x y 
                                                 \key frame modifier-state event}

Given an input context \arg{input-context}, an output recording window stream
\arg{window}, $x$ and $y$ positions \arg{x} and \arg{y}, returns the innermost
presentation whose sensitivity region contains \arg{x} and \arg{y} that matches
the innermost input context, using the translator matching algorithm described
below.  If there is no such presentation, this function will return \cl{nil}.

\arg{event} and \arg{modifier-state} are a pointer button event and modifier
state (see \cl{event-modifier-key-state}).  \arg{event} defaults to \cl{nil},
and \arg{modifier-state} defaults to the current modifier state for
\arg{window}.  Only one of \arg{event} or \arg{modifier-state} may be supplied;
it is unspecified what will happen if both are supplied.

\arg{frame} defaults to the current frame, \cl{*application-frame*}.

The default method for \cl{frame-find-innermost-applicable-presentation} will
call this function.


\Defun {throw-highlighted-presentation} {presentation input-context button-press-event}

Given a \term{presentation} \arg{presentation}, input context
\arg{input-context}, and a button press event (which contains the window,
pointer, $x$ and $y$ position of the pointer within the window, the button
pressed, and the modifier state), find the translator that matches the innermost
presentation in the innermost input context, then call the translator to produce
an object and a presentation type.  Finally, the matching input context that was
established by \cl{with-input-context} will be terminated.

Note that it is possible that more than one translator having the same gesture
may be applicable to \arg{presentation} in the specified input context.  In this
case, the translator having the highest priority will be chosen.  If there is
more than one having the same priority, it is unspecified what translator will
be chosen.


\Defun {highlight-applicable-presentation} {frame stream input-context
                                            \optional prefer-pointer-window}

This is the core of the ``input wait'' handler used by \cl{with-input-context}
on behalf of the \term{application frame} \arg{frame}.  It is responsible for
locating the innermost applicable presentation on \arg{stream} in the input
context \arg{input-context}, unhighlighting presentations that are not
applicable, and highlighting the presentation that is applicable.  Typically on
entry to \cl{highlight-applicable-presentation}, \arg{input-context} will be the
value of \cl{*input-context*} and \arg{frame} will be the value of
\cl{*application-frame*}.

If \arg{prefer-pointer-window} is \term{true} (the default), CLIM will highlight
the applicable presentation on the same window that the pointer is located over.
Otherwise, CLIM will highlight an applicable presentation on \arg{stream}.

{\bf Implementation note:} This will probably use
\cl{frame-find-innermost-applicable-presentation-at-position} to locate the
innermost presentation, and \cl{unhighlight-highlighted-presentation} and
\cl{set-highlighted-presentation} to unhighlight and highlight presentations.


\Defun {set-highlighted-presentation} {stream presentation \optional prefer-pointer-window}

Highlights the \term{presentation} \arg{presentation} on \arg{stream}.  This
must call \cl{highlight-presentation} methods if that is appropriate.

\arg{prefer-pointer-window} is as for \cl{highlight-applicable-presentation}.


\Defun {unhighlight-highlighted-presentation} {stream \optional prefer-pointer-window}

Unhighlights any highlighted presentations on \arg{stream}.

\arg{prefer-pointer-window} is as for \cl{highlight-applicable-presentation}.


\subsection {Translator Applicability}

The top-level ``input wait'', which is what you are in when inside of a
\cl{with-input-context}, is responsible for determining what translators are
applicable to which presentations in a given input context.  This loop both
provides feedback in the form of highlighting sensitive presentation, and is
responsible for calling the applicable translator when the user presses a
pointer button.

{\bf Implementation note:} \cl{with-input-context} uses
\cl{frame-find-innermost-applicable-presentation-at-position} (via
\cl{highlight-applicable-presentation}) as its ``input wait'' handler, and
\cl{frame-input-context-button-press-handler} as its button press ``event
handler''.

Given a presentation, an input context established by \cl{with-input-context},
and an event corresponding to a user gesture, translator matching proceeds as
follows. 

The set of candidate translators is initially those translators accessible in
the command table in use by the current application.  A translator is said to
``match'' if all of the following are true (in this order):

\begin{enumerate}
\item The presentation's type is \cl{presentation-subtypep} of the translator's
\arg{from-type}, ignoring type parameters.

\item The translator's \arg{to-type} is \cl{presentation-subtypep} of the input
context type, ignoring type parameters.

\item The translator's gesture is either \cl{t}, or matches the event
corresponding to the user's gesture.

\item If there are parameters in the \arg{from-type}, the presentation's object
must be \cl{presentation-typep} of the \arg{from-type}.

\item The translator's tester returned \term{true}.  If there is no tester, the
translator behaves as though there is a tester that always returns \term{true}.

\item If there are parameters in the input context type and the tester is not
declared to be definitive, the value returned by body of the translator must be
\cl{presentation-typep} of the context type.
\end{enumerate}

Note that the type parameters from the presentation's type have no effect on
translator lookup.

\cl{find-presentation-translator} is responsible for the first two steps of the
matching algorithm, and \cl{test-presentation-translator} is responsible for the
remaining steps.

When a single translator is being chosen (such as is done by
\cl{throw-highlighted-presentation}), it is possible that more than one
translator having the same gesture may be applicable to the presentation in the
specified input context.  In this case, the translator having the highest
priority will be chosen.  If there is more than one having the same priority, it
is unspecified what translator will be chosen.

The matching algorithm is somewhat more complicated in face of nested
presentations and nested input contexts.  In this case, the applicable
presentation is the {\sl smallest} presentation that matches the {\sl innermost}
input context.

Sometimes there may be nested presentations that have exactly the same bounding
rectangle.  In this case, it is not possible for a user to unambiguously point
to just one of the nested presentations.  Therefore, when CLIM has located the
innermost applicable presentation in the innermost input context, it must then
search for outer presentations having exactly the same bounding rectangle,
checking to see if there are any applicable translators for those presentations.
If there are multiple applicable translators, the one having the highest
priority is chosen.  \cl{find-applicable-translators},
\cl{call-presentation-menu}, \cl{throw-highlighted-presentation}, and the
computation of pointer documentation must all take this situation into account.

The translators are searched in the order that they are returned by
\cl{find-presentation-translators}.  The rules for the ordering of the
translators are described under that function.


\section {Standard Presentation Types}

The following sections document the presentation types supplied by CLIM.  Any
presentation type with the same name as a Common Lisp type accepts the same
parameters as the Common Lisp type (and additional parameters in a few cases).


\subsection {Basic Presentation Types}

\Defptype {t} {}

The supertype of all other presentation types.

\Defptype {nil} {}

The subtype of all other presentation types.  This has no printed
representation, and it useful only in writing ``context independent''
translators, that is, translators whose \arg{to-type} is \cl{nil}.

\Defptype {null} {}

The type that represents ``nothing''.  The single object associated with this
type is \cl{nil}, and its printed representation is ``None''.

\Defptype {boolean} {}

The type that represents \term{true} or \term{false}.  The printed
representation is ``Yes'' or ``No'', respectively.

\Defptype {symbol} {}

The type that represents a symbol.

\Defptype {keyword} {}

The type that represents a symbol in the keyword package.  It is a subtype of
\cl{symbol}.

\Defptype {blank-area} {}

The type that represents all the places in a window where there is no
presentation that is applicable in the current input context.  CLIM provides a
single ``null presentation'' as the object associated with this type.

\Defconst {*null-presentation*}

The null presentation, which occupies all parts of a window in which there are
no applicable presentations.  This will have a presentation type of \cl{blank-area}.


\subsection {Numeric Presentation Types}

\Defptype {number} {}

The type that represents a general number.  It is the supertype of all the
number types.

\Defptype {complex} {\optional type}

The type that represents a complex number.  It is a subtype of \cl{number}.

The components of the complex number are of type \arg{type}, which must be
\cl{real} or a subtype of \cl{real}.

\Defptype {real} {\optional low high}

The type that represents either a ratio, an integer, or a floating point number
between \arg{low} and \arg{high}.  \arg{low} and \arg{high} can be inclusive or
exclusive, as in Common Lisp type specifiers.  Options to this type are
\arg{base} (default 10) and \arg{radix} (default \cl{nil}).  \cl{real} is a
subtype of \cl{number}.

\Defptype {rational} {\optional low high}

The type that represents either a ratio or an integer between \arg{low} and
\arg{high}.  Options to this type are \arg{base} and \arg{radix}.  \cl{rational}
is a subtype of \cl{real}.

\Defptype {integer} {\optional low high}

The type that represents an integer between \arg{low} and \arg{high}.  Options
to this type are \arg{base} and \arg{radix}.  \cl{integer} is a subtype of
\cl{rational}.

\Defptype {ratio} {\optional low high}

The type that represents a ratio between \arg{low} and \arg{high}.  Options to
this type are \arg{base} and \arg{radix}.  \cl{ratio} is a subtype of
\cl{rational}.

\Defptype {float} {\optional low high}

The type that represents a floating point number between \arg{low} and
\arg{high}.  \cl{float} is a subtype of \cl{number}.


\subsection {Character and String Presentation Types}

\Defptype {character} {}

The type that represents a character object.

\Defptype {string} {\optional length}

The type that represents a string.  If \arg{length} is supplied, the string
must contain exactly that many characters.


\subsection {Pathname Presentation Type}

\Defptype {pathname} { }

The type that represents a pathname.  The options are \arg{default-version},
which defaults to \cl{:newest}, \arg{default-type}, which defaults to \cl{nil},
and \arg{merge-default}, which defaults to \term{true}.  If \arg{merge-default}
is \term{false}, \cl{accept} returns the exact pathname that was entered,
otherwise \cl{accept} merges against the default and \arg{default-version}.  If
no default is supplied, it defaults to \cl{*default-pathname-defaults*}.  The
\cl{pathname} type should have a default preprocessor that merges the options
into the default.


\subsection {``One-of'' and ``Some-of'' Presentation Types}

\Defptype {completion} {sequence \key test value-key}

The type that selects one from a finite set of possibilities, with
``completion'' of partial inputs.  The member types below, \cl{token-or-type},
and \cl{null-or-type} are implemented in terms of the \cl{completion} type.

\arg{sequence} is a list or vector whose elements are the possibilities.  Each
possibility has a printed representation, called its name, and an internal
representation, called its value.  \cl{accept} reads a name and returns a value.
\cl{present} is given a value and outputs a name.

\arg{test} is a function that compares two values for equality.  The default is
\cl{eql}.

\arg{value-key} is a function that returns a value given an element of
\arg{sequence}.  The default is \cl{identity}.

The following presentation type options are available:

\arg{name-key} is a function that returns a name, as a string, given an element
of \arg{sequence}.  The default is a function that behaves as follows: \\
\begin{tabular}{rcl}
string & $\Rightarrow$ & the string \\
null & $\Rightarrow$ & \cl{"NIL"} \\
cons & $\Rightarrow$ & \cl{string} of the \cl{car} \\
symbol & $\Rightarrow$ & \cl{string-capitalize} of its name \\
otherwise & $\Rightarrow$ & \cl{princ-to-string} of it
\end{tabular}

\arg{documentation-key} is a function that returns either \cl{nil} or a
descriptive string, given an element of \arg{sequence}.  The default always
returns \cl{nil}.

\arg{test}, \arg{value-key}, \arg{name-key}, and \arg{documentation-key} must
have indefinite extent.

\arg{partial-completers} is a possibly-empty list of characters that delimit
portions of a name that can be completed separately.  The default is a list of
one character, \verb+#\Space+.


\DefptypeAbbrev {member} {\rest elements}

The type that specifies one of \arg{elements}.  The options are the same as for
\cl{completion}.


\DefptypeAbbrev {member-sequence} {sequence \key test}

Like \cl{member}, except that the set of possibilities is the sequence
\arg{sequence}.  The parameter \arg{test} and the options are the same as for
\cl{completion}.


\DefptypeAbbrev {member-alist} {alist \key test}

Like \cl{member}, except that the set of possibilities is the alist \arg{alist}.
Each element of \arg{alist} is either an atom as in \cl{member-sequence} or a
list whose \cl{car} is the name of that possibility and whose \cl{cdr} is one of
the following:

\begin{itemize}
\item The value (which must not be a cons)

\item A list of one element, the value

\item A property list that can contain the following properties:
  \begin{itemize}
  \item \cl{:value}---the value
  \item \cl{:documentation}---a descriptive string
  \end{itemize}
\end{itemize}

The \arg{test} parameter and the options are the same as for \cl{completion}
except that \arg{value-key} and \arg{documentation-key} default to functions
that support the specified alist format.


\Defptype {subset-completion} {sequence \key test value-key}

The type that selects one or more from a finite set of possibilities, with
``completion'' of partial inputs.  The parameters and options are the same as
for \cl{completion}, plus the additional options \arg{separator} and
\arg{echo-space}, which are as for the \cl{sequence} type.  The subset types
below are implemented in terms of the \cl{subset-completion} type.


\DefptypeAbbrev {subset} {\rest elements}

The type that specifies a subset of \arg{elements}.  Values of this type are
lists of zero or more values chosen from the possibilities in \arg{elements}.
The printed representation is the names of the elements separated by commas.
The options are the same as for \cl{completion}.


\DefptypeAbbrev {subset-sequence} {sequence \key test}

Like \cl{subset}, except that the set of possibilities is the sequence
\arg{sequence}.  The parameter \arg{test} and the options are the same as for
\cl{completion}.


\DefptypeAbbrev {subset-alist} {alist \key test}

Like \cl{subset}, except that the set of possibilities, the parameters, and the
options are as for \cl{member-alist}.


\subsection {Sequence Presentation Types}

\Defptype {sequence} {type}

The type that represents a sequence of elements of type \arg{type}.  \arg{type}
can be a presentation type abbreviation.  The printed representation of a
\cl{sequence} type is the elements separated by commas.  It is unspecified
whether \cl{accept} returns a list or a vector.

The options to this type are \arg{separator} and \arg{echo-space}.
\arg{separator} is used to specify a character that will act as the separator
between elements of the sequence; the default is the comma character \verb+#\,+.
\arg{echo-space} must be \term{true} or \term{false}; when it is \term{true}
(the default) a space will be automatically inserted into the input buffer when
the user types a separator character.


\Defptype {sequence-enumerated} {\rest types}

\cl{sequence-enumerated} is like \cl{sequence}, except that the type of each
element in the sequence is individually specified.  The elements of \arg{types}
can be presentation type abbreviations.  It is unspecified whether \cl{accept}
returns a list or a vector.

The options to this type are \arg{separator} and \arg{echo-space}, which are as
for the \cl{sequence} type.


\subsection {``Meta'' Presentation Types}

\Defptype {or} {\rest types}

The type that is used to specify one of several types, for example, \cl{(or
(member :all :none) integer)}.  The elements of \arg{types} can be presentation
type abbreviations.  \cl{accept} returns one of the possible types as its second
value, not the original \cl{or} presentation type specifier.

The \cl{accept} method for \cl{or} could be implemented by iteratively calling
\cl{accept} on each of the presentation types in \arg{types}.  It would
establish a condition handler for \cl{parse-error}, call \cl{accept} on one of
the types and return the result if no condition was signalled.  If a
\cl{parse-error} is signalled, the \cl{accept} method for \cl{or} would call
\cl{accept} on the next type.  When there are no more types, the \cl{accept}
method for \cl{or} would itself signal a \cl{parse-error}.


\Defptype {and} {\rest types}

The type that is used for ``multiple inheritance''.  \cl{and} is frequently used
in conjunction with \cl{satisfies}, for example, \cl{(and integer (satisfies
oddp))}.  The elements of \arg{types} can be presentation type abbreviations.

The \cl{and} type has special syntax that supports the two ``predicates'',
\cl{satisfies} and \cl{not}.  \cl{satisfies} and \cl{not} cannot stand alone as
presentation types and cannot be first in \arg{types}.  \cl{not} can surround
either \cl{satisfies} or a presentation type.

The first type in \arg{types} is the type whose methods will be used during
calls to \cl{accept} and \cl{present}.


\subsection {Compound Presentation Types}

\DefptypeAbbrev {token-or-type} {tokens type}

A compound type that is used to select one of a set of special tokens, or an
object of type \arg{type}.  \arg{tokens} is anything that can be used as the
\arg{sequence} parameter to \cl{member-alist}; typically it is a list of
symbols.


\DefptypeAbbrev {null-or-type} {type}

A compound type that is used to select \cl{nil}, whose printed representation is
the special token ``None'', or an object of type \arg{type}.


\DefptypeAbbrev {type-or-string} {type}

A compound type that is used to select an object of type \arg{type} or an
arbitrary string, for example, \cl{(type-or-string integer)}.  Any input that
\cl{accept} cannot parse as the representation of an object of type \arg{type}
is returned as a string.


\subsection {Lisp Expression Presentation Types}

\Defptype {expression} {}

The type used to represent any Lisp object.  The standard \cl{print} and
\cl{read} functions produce and accept the textual view of this type.

If a presentation history is maintained for the \cl{expression} presentation
type, it should be maintained separately for each instance of an application
frame.

\Defptype {form} {}

The type used to represent a Lisp form.  This is a subtype of \cl{expression}
and is equivalent except that some presentation translators produce \cl{quote}
forms.
