\chapter{CLIM demos and applications}

\section{Running the demos}

The McCLIM source distribution comes with a number of demos and
applications.  They are intended to showcase specific CLIM features,
demonstrate programming techniques or provide useful tools.

These demos and applications are available in the \texttt{Examples}
and \texttt{Apps} subdirectories of the source tree's root directory.
Instructions for compiling, loading and running some of the demos are
included in the file \texttt{INSTALL} with the McCLIM installation
instructions for your Common Lisp implementation.

Below is a complete list of the McCLIM demos and applications, sorted in
alphabetical order.  Each entry provides a short description of what the
program does, with instructions for compiling and running it if not
mentioned in the general installation instructions.

Demos are meant to be run after loading the \texttt{clim-examples}
system from the frame created with \texttt{(clim-demo:demodemo)}.

The easiest way to try this is to use \emph{Quicklisp} library
manager. Assuming that this is already setup, trying out the demos is
straightforward:

\begin{verbatim}
(ql:quickload 'clim-examples)
(clim-demo:demodemo)
\end{verbatim}


Alternatively, for the more courageous (which requires separately
downloading dependencies and setting a local repository),
\texttt{asdf} also works well starting from the McCLIM source code
directory.

\begin{verbatim}
(asdf:load-system 'clim-examples)
(clim-demo:demodemo)
\end{verbatim}

Available demos and tests are defined in the following files:

\begin{itemize}
\item Examples/demodemo.lisp

  Demonstrates different pane types and other tests.

\item Examples/clim-fig.lisp

  Simple paint program.

\item Examples/calculator.lisp

  Simple desk calculator.

\item Examples/method-browser.lisp

  Example of how to write a CLIM application with a ``normal'' GUI,
  where ``normal'' is a completely event driven app built using
  gadgets and not using the command-oriented framework.

\item Examples/address-book.lisp

  Simple address book.

\item Examples/puzzle.lisp

  Simple puzzle game.

\item Examples/colorslider.lisp

  Interactive color editor.

\item Examples/town-example.lisp

  ``Large Cities of Germany'' application example by Max-Gerd Retzlaff.

\item Examples/logic-cube.lisp

  Software-rendered 3d logic cube game. Shows how the transformations
  work and how to implement custom handle-repaint methods.

\item Examples/menutest.lisp

  Displays a window with a simple menu bar.

\item Examples/gadget-test.lisp

  Displays a window with various gadgets.

\item Examples/dragndrop.lisp

  Example of ``Drag and Drop'' functionality.

\item Examples/dragndrop-translator.lisp

  Another example of ``Drag and Drop'' functionality (with colors!).

\item Examples/draggable-graph.lisp

  Demo of draggable graph nodes.

\item Examples/image-viewer.lisp

  A simple program for displaying images of formats known to McCLIM.

\item Examples/font-selection.lisp

  A font selection dialog.

\item Examples/tabdemo.lisp

  A tab layout demo (McCLIM extension).

\item Examples/postscript-test.lisp

  Displays text and graphics to a PostScript file.  Run it with:

\begin{verbatim}
(clim-demo::postscript-test)
\end{verbatim}

  The resulting file \texttt{ps-test.ps} is generated in the current
  directory and can be displayed by a PostScript viewer such as
  \texttt{gv} on Unix-like systems.

\item Examples/presentation-test.lisp

  Displays an interactive window in which you type numbers that are
  successively added.  When a number is expected as input, you can
  either type it at the keyboard, or click on a previously entered
  number. Labeled ``Summation''.

\item Examples/sliderdemo.lisp

  Apparently a calculator demo (see above). Labeled ``Slider demo''.

\item Examples/stream-test.lisp

  Interactive command processor that echoes its input in
  \texttt{*trace-output*}.

\end{itemize}

The following programs are currently \textbf{known not to work}:
\begin{itemize}
\item
 \texttt{Examples/gadget-test-kr.lisp}
\item
 \texttt{Examples/traffic-lights.lisp}
\end{itemize}

\section{Applications}

\begin{itemize}
\item Apps/Listener

  CLIM-enabled Lisp listener. System name is
  \texttt{clim-listener}. See instructions in
  \texttt{Apps/Listener/README} for more information.

\item Apps/Inspector

  CLIM-enabled Lisp inspector. System name is \texttt{clouseau}. See
  instructions in \texttt{Apps/Inspector/INSTALL} for more
  information..

\item Apps/Debugger

  Common Lisp debugger implemented in McCLIM. It uses the portable
  debugger interface developed for the Slime project. Application has
  some quirks and requires work. System name is
  \texttt{clim-debugger}.

\item Apps/Functional-Geometry

  Peter Henderson idea, see http://www.ecs.soton.ac.uk/~ph/funcgeo.pdf
  and http://www.ecs.soton.ac.uk/~ph/papers/funcgeo2.pdf implemented
  in Lisp by Frank Buss. CLIM Listener interface by Rainer
  Joswig. System name is \texttt{functional-geometry}.

\begin{verbatim}
(functional-geometry:run-functional-geometry)
(clim-plot *fishes*) ; from a listener
\end{verbatim}

\item Apps/Scigraph

  Scigraph Scientific Graphing Package. See the compilation and
  execution instructions in \texttt{Apps/Scigraph/README}.
\end{itemize}

