\chapter{Fonts and Extended Text Styles}

\section{Extended Text Styles}

McCLIM extends the legal values for the \texttt{family} and \texttt{face}
arguments to \texttt{make-text-style} to include strings (in additional to
the portable keyword symbols), as permitted by the CLIM spec, section
11.1.

Each backend defines its own specific syntax for these family and face
names.

The CLX backend maps the text style family to the X font's
\emph{foundry} and \emph{family} values, separated by a dash.  The
face is mapped to \emph{weight} and \emph{slant} in the same way.  For
example, the following form creates a text style for
\emph{-misc-fixed-bold-r-*-*-18-*-*-*-*-*-*-*}:

\begin{verbatim}
(make-text-style "misc-fixed" "bold-r" 18)
\end{verbatim}

In the GTK backend, the text style family and face are used directly
as the Pango font family and face name.  Please refer to Pango
documentation for details on the syntax of face names.  Example:

\begin{verbatim}
(make-text-style "Bitstream Vera Sans" "Bold Oblique" 54)
\end{verbatim}

\section{Listing Fonts}

McCLIM's font listing functions allow applications to list all
available fonts available on a \texttt{port} and create text style
instances for them.

Example:

\begin{verbatim}
* (find "Bitstream Vera Sans Mono"
        (clim-extensions:port-all-font-families (clim:find-port))
        :key #'clim-extensions:font-family-name
        :test #'equal)
#<CLIM-GTKAIRO::PANGO-FONT-FAMILY Bitstream Vera Sans Mono>

* (clim-extensions:font-family-all-faces *)
(#<CLIM-GTKAIRO::PANGO-FONT-FACE Bitstream Vera Sans Mono, Bold>
 #<CLIM-GTKAIRO::PANGO-FONT-FACE Bitstream Vera Sans Mono, Bold Oblique>
 #<CLIM-GTKAIRO::PANGO-FONT-FACE Bitstream Vera Sans Mono, Oblique>
 #<CLIM-GTKAIRO::PANGO-FONT-FACE Bitstream Vera Sans Mono, Roman>)

* (clim-extensions:font-face-scalable-p (car *))
T

* (clim-extensions:font-face-text-style (car **) 50)
#<CLIM:STANDARD-TEXT-STYLE "Bitstream Vera Sans Mono" "Bold" 50>
\end{verbatim}

\Defclass {font-family}

The protocol class for font families.  Each backend defines a subclass
of font-family and implements its accessors.  Font family instances
are never created by user code.  Use port-all-font-families to list
all instances available on a port.

\Defclass {font-face}

The protocol class for font faces Each backend defines a subclass of
font-face and implements its accessors.  Font face instances are never
created by user code.  Use font-family-all-faces to list all faces of
a font family.

\Defgeneric {port-all-font-families} {port \key invalidate-cache \allow}

Returns the list of all \texttt{font-family} instances known by
\texttt{port}.  With \texttt{invalidate-cache}, cached font family
information is discarded, if any.

\Defgeneric {font-family-name} {font-family}

Return the font family's name.  This name is meant for user display,
and does not, at the time of this writing, necessarily the same string
used as the text style family for this port.

\Defgeneric {font-family-port} {font-family}

Return the port this font family belongs to.

\Defgeneric {font-family-all-faces} {font-family}

Return the list of all font-face instances for this family.

\Defgeneric {font-face-name} {font-face}

Return the font face's name.  This name is meant for user display, and
does not, at the time of this writing, necessarily the same string
used as the text style face for this port.

\Defgeneric {font-face-family} {font-face}

Return the font family this face belongs to.

\Defgeneric {font-face-all-sizes} {font-face}

Return the list of all font sizes known to be valid for this font, if
the font is restricted to particular sizes.  For scalable fonts,
arbitrary sizes will work, and this list represents only a subset of
the valid sizes.  See font-face-scalable-p.

\Defgeneric {font-face-scalable-p} {font-face}

Return true if this font is scalable, as opposed to a bitmap font.
For a scalable font, arbitrary font sizes are expected to work.

\Defgeneric {font-face-text-style} {font-face \optional size}

Return an extended text style describing this font face in the specified
size.  If size is nil, the resulting text style does not specify a size.
