\chapter{Menu bar}
\index{menu-bar}

Menu bar has become essential part of every GUI system, including
McClim. Ideally, McClim should try to use the menu bar provided by
host window system via McClim backends, but the current
\texttt{clx-backend} doesn't supports native menu bars. That's why it
has some quirks of its own, like you need to keep mouse button pressed
while accessing the sub-menus.

\section{Creating Menu bar}
\label{creating-menu-bar}

McCLIM makes creating menu bar quite easy.

\begin{verbatim}
        (clim:define-application-frame foo ()
        ...
        (:menu-bar t)
        ...)
\end{verbatim}

Options for \texttt{:menu-bar} can be
\begin{itemize}
\item
  \texttt{t} (default) : McClim will provide the menu bar. Later, when
  you start defining commands, you can provide a \texttt{(:menu t)}
  argument to command definition that will add this command to menu
  bar.
\item
  \texttt{nil} : McClim won’t provide the menu bar.
\item
  \texttt{command-table} : If you provide a named command table as
  argument, that command table is used to provide the menu bar(See
  Chapter \ref{using-command-tables} on command tables).
\end{itemize}

To add a sub-menu to menu bar, You need to change the type of
menu-item from \texttt{:command} to \texttt{:menu} (which requires
another \texttt{command-table} as argument) (See
\refSec{modifying-menu-items}).

\section{Modifying Menu bar}

Menu bar can be changed anytime by changing \texttt{command-table}
associated with the current \texttt{frame}. For example:
\begin{verbatim}
        (setf (frame-command-table *application-frame*)
              new-command-table)
\end{verbatim}
changes menu bar of \texttt{*application-frame*} by replacing current
\texttt{command-table} (accessible by \texttt{frame-command-table}
function) with \texttt{new-command-table}.

\subsection{Modifying menu items of command table}
\label{modifying-menu-items}

Menu items can be added to command table with following function:

\Defun {add-menu-item-to-command-table} {command-table string type value
  \\\key documentation (after ':end) keystroke text-style (errorp t)}

Arguments to this function are:
\begin{itemize}
\item
  \texttt{command-table} : Command table to which we want to add the menu item.
\item
  \texttt{string} : name of the menu item as it will appear on the
  menu bar. Its character case is ignored e.g. you may give it
  \texttt{file} or \texttt{FILE} but it will appear as \texttt{File}.
\item
  \texttt{type} and \texttt{value} : type could be one of
  \texttt{:command},\texttt{:function},\texttt{:menu} and
  \texttt{:divider}. value of \texttt{value} depends on
  \texttt{type}. So when given the
  \begin{itemize}
  \item
    \texttt{:command} : \texttt{value} must be a command, a cons of
    command name and it's arguments. if you omit the arguments McCLIM
    will prompt for arguments.
  \item
    \texttt{:function} : \texttt{value} must be a function having
    indefinite extent that, when called, returns a command. Function
    must accept two arguments, the gesture (keyboard or mouse press
    event) and a ``numeric argument''.
  \item
    \texttt{:menu} : \texttt{value} must be another command
    table. This type is used to add sub-menus to the menu.
  \item
    \texttt{:divider} : \texttt{value} is ignored, and \texttt{string}
    is used as a divider string. Using ``|'' as string will make it
    obvious to users that it is a divider.
  \end{itemize}
\item
  \texttt{documentation} : You can provide the documentation (for
  non-obvious menu items) which will be displayed on
  pointer-documentation pane (if you have one).
\item
  \texttt{after} (default \texttt{:end}) : This determines where item
  will be inserted in the menu. The default is to add it to the
  end. Other values could be \texttt{:start}, \texttt{:sort}(add in
  alphabetical order) or \texttt{string} which is name of existing
  menu-item to add after it.
\item
  \texttt{keystroke} : If keystroke is supplied, it will be added to
  comand tables keystroke accelerator table. Value must be a keyboard
  gesture name e.g. (:s :control) for Control + s.
\item
  \texttt{text-style} : This is either a text style spec or nil. It is
  used to indicate that the command menu item should be drawn with the
  supplied text style in command menus.
\item
  \texttt{error-p} : If this is \texttt{t} and the item already exists
  in the menu, it will signal a \texttt{command-already-present}
  error. If \texttt{nil}, it will first remove the existing item and
  add the new item to the command-table.
\end{itemize}

To remove items from command table, the following function is used:

\Defun {remove-menu-item-from-command-table} {command-table string \key (errorp t)}

where \texttt{command-table} is command-table-designator and
\texttt{string} is the menu item's name (it is case-insensitive). You can
provide \texttt{:error-p nil} to suppress the error if item is not in
the command-table.

Note that both of the above functions \texttt{\emph{do not}}
automatically update the menu bar. For that you need to replace the
existing \texttt{frame-command-table} with the modified command table
using \texttt{setf}. The typical way to do this is to use \texttt{let}
to create a copy of \texttt{frame-command-table}, modify it and at the
end call \texttt{setf} to replace the original.
