\chapter{Raster Image backend}
Raster image backend includes a medium that supports:
\begin{itemize}
\item CLIM's medium protocol, 
\item CLIM's output stream protocol, and
\item CLIM's Pixmap protocol.
\end{itemize}

Package \texttt{mcclim-raster-image} exports the following macros:

\Defmacro {mcclim-render:with-output-to-raster-image-stream} {(stream-var file-stream format \rest options) \body body}
\Defmacro {mcclim-render:with-output-to-rgb-pattern} {(stream-var image \rest options) \body body}

Within \arg{body}, \arg{stream-var} is bound to a stream that produces
a raster image.  This stream is suitable as a stream or medium argument to any
CLIM output utility, such as \cl{draw-line*} or \cl{write-string}.  

The value of \arg{options} is a list consisting of alternating keyword and value
pairs.  These are the supported keywords:
\begin{itemize}
\item \cl{:width}---specifies the width of the image. Its default value is 1000.
\item \cl{:height}---specifies the height of the image. Its default value is 1000. 
\end{itemize}

\Defmacro {mcclim-render:with-output-to-raster-image-stream} {(stream-var file-stream format \rest options) \body body}

An image describing the output to the \arg{stream-var} stream will be
written to the stream \arg{file-stream} using the format
\arg{format}. \arg{format} is a symbol that names the type of the
image. Valid values are \cl{:png}, \cl{:jpg}, \cl{:jpeg}, \cl{tiff},
\cl{tif}, \cl{gif}, \cl{pbm}, \cl{pgm}, and \cl{ppm}. Its default
value is \cl{:png}.

\Defmacro {mcclim-render:with-output-to-rgb-pattern} {(stream-var image \rest options) \body body}

An image describing the output to the \arg{stream-var} stream will be
returned as an rgb-pattern (of class \cl{climi::rgb-pattern}).


To run an example:
\begin{verbatim}
(ql:quickload :clim-examples)
(load "Examples/drawing-tests")
(clim-demo::run-drawing-tests)
\end{verbatim}




